---
title: "Cephalic furrow depth in wildtype/heterozygote cauterizations"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: April 06, 2021
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/germband-cauterizations")
library(ggplot2)
library(ggbeeswarm)
```

## Load measurements

```{r load}
# Load raw ROI measurements from ImageJ
results <- read.csv("1-wt/Lengths.csv")

# Discard cauterized embryo where the GB wasn't blocked from extending
results <- results[results$Embryo != "2021-03-24_E5",]

# Write data frame to file
write.csv(results, file = "depth.csv", row.names = FALSE)
```

## Print summary

```{r summary}
print_stats <- function(dataframe, stock, sample) {
  filter_df <- (dataframe$Stock == stock & dataframe$Sample == sample)
  df <- dataframe[filter_df,]
  df <- droplevels(df)

  # Define row
  row <- c(stock,
           sample,
           mean(df$Length),
           sd(df$Length),
           length(df$Length),
           length(unique(df$Embryo))
           )
  
  # Summarize data
  print(sprintf("%s %s / area=%.3f±%.3f x 10^3 / n_folds=%i / n_embryos=%i",
                stock, sample,
                mean(df$Length),
                sd(df$Length),
                length(df$Length),
                length(unique(df$Embryo))
                ))
  return(row)
}

# Save stats to data frame
stats <- data.frame()

# Bind row by row
stats <- rbind(stats, print_stats(results, "wt", "control"))
stats <- rbind(stats, print_stats(results, "wt", "cauterized"))
stats <- rbind(stats, print_stats(results, "btd", "control"))
stats <- rbind(stats, print_stats(results, "btd", "cauterized"))
stats <- rbind(stats, print_stats(results, "eve", "control"))
stats <- rbind(stats, print_stats(results, "eve", "cauterized"))

# Add column names
colnames(stats) <- c("stock", "sample", "mean", "sd", "n_folds", "n_embryos")

# Convert columns to numeric
stats[, 3:5] <- apply(stats[, 3:5], 2, as.numeric)

# Round up
stats[, 3:4] <- round(stats[, 3:4], digits=1)

# Write table to file
write.csv(format(stats, digits=2), file = "wt-stats.csv", row.names = FALSE)
print(stats)
```

## Maximum depth of individual folds

```{r depth-ind}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(results, aes(factor(Sample, level = c("control", "cauterized")), Length, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  labs(x = "sample", y = "depth (µm)") +
  facet_wrap(~Stock) +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(size = 45, face = "italic"),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-length.png", width = 7, height = 4)
ggsave("4-plots/wt-length.svg", width = 7, height = 4)
```

## Maximum depth of individual folds, faceted by condition

```{r depth-ind-condition}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(results, aes(factor(Sample, level = c("control", "cauterized")), Length, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  labs(x = "sample", y = "depth (µm)") +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(size = 45, face = "italic"),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-length-condition.png", width = 4, height = 4)
ggsave("4-plots/wt-length-condition.svg", width = 4, height = 4)
```

## Mean furrow length by embryo

To attempt to normalize the furrow length per embryo, since their length does not
seem to be independent of each other, I'll take the length mean per embryo.

```{r mean-length}
# Mean of CF length per embryo
embryo_cf_mean <- aggregate(results$Length, list(results$Embryo), mean)
colnames(embryo_cf_mean) <- c("Embryo", "Mean")

# Standard deviation of CF length per embryo
embryo_cf_sd <- aggregate(results$Length, list(results$Embryo), sd)
colnames(embryo_cf_sd) <- c("Embryo", "SD")

# Get sample information
samples <- unique(results[, c("Stock", "Embryo", "Sample")])

# Merge datasets
means <- merge(samples, embryo_cf_mean, by = "Embryo", all.x = TRUE)
means <- merge(means, embryo_cf_sd, by = "Embryo", all.x = TRUE)

# Write to file
write.csv(means, file = "means.csv", row.names = FALSE)

# Summary by sample
by(means[means$Sample == "control", c("Mean", "SD")], means[means$Sample == "control",]$Stock, summary)
print("---")
by(means[means$Sample == "cauterized", c("Mean", "SD")], means[means$Sample == "cauterized",]$Stock, summary)
print("---")
# Summary by condition
by(means[, c("Mean", "SD")], means$Sample, summary)
print("---")
print("Calculating difference between non-cauterized and cauterized groups")
# Statistical testing
diff_micron <- mean(means[means$Sample == "control",]$Mean) - mean(means[means$Sample == "cauterized",]$Mean)
diff_pct <- diff_micron / mean(means[means$Sample == "cauterized",]$Mean) * 100
print(sprintf("%s microns shallower", diff_micron))
print(sprintf("%s percent shallower", diff_pct))
```

## Maximum depth of mean length by embryo

Due to the low n, we pooled the different genotypes together. There's a small
difference in length of about 15% (8 microns). But this needs more data to be
conclusive.

```{r depth-mean}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(means, aes(factor(Sample, level = c("control", "cauterized")), Mean, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  labs(x = "condition", y = "depth (µm)") +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 25),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-mean-length.png", width = 4, height = 4)
ggsave("4-plots/wt-mean-length.svg", width = 4, height = 4)
```

```{r depth-mean-facet}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(means[means$Stock != "eve",], aes(factor(Sample, level = c("control", "cauterized")), Mean, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_pointrange(aes(ymin=Mean-SD, ymax=Mean+SD), alpha = 0.8, size = 0.5, show.legend = FALSE, position = position_quasirandom(width = 0.1)) +
  facet_wrap(~Stock) +
  labs(x = "condition", y = "depth (µm)") +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(size = 45, face = "italic"),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-mean-length-facet.png", width = 4, height = 4)
ggsave("4-plots/wt-mean-length-facet.svg", width = 4, height = 4)
```


## Statistical tests

Sample size is small for normality test, defaulting to Mann-Whitney. Using the embryo mean length as measurements of left and right side might not be independent. Choosing one-sided to test if cauterized is smaller than control.

```{r statistical-tests}
# Create data frame for tests
tests <- data.frame()

# Compare means between combined conditions as in the main plot
ctrl <- means[means$Sample == "control", "Mean"]
caut <- means[means$Sample == "cauterized", "Mean"]

# Control vs Cauterized
contrast <- "ctrl vs caut"
test <- wilcox.test(caut, ctrl, alternative = "less", paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
mean_ctrl <- mean(ctrl)
mean_caut <- mean(caut)
delta <- mean_ctrl - mean_caut

test_row <- c(contrast = contrast,
              data = "means",
              sample = "pooled",
              mean_ctrl = mean_ctrl,
              mean_caut = mean_caut,
              delta = sprintf("%.1f", delta),
              significant = pvalue < 0.05,
              pvalue = sprintf("%.9f", pvalue),
              statistic = sprintf("%s=%s", names(statistic), statistic),
              paired = FALSE)
tests <- rbind(tests, as.data.frame(t(test_row)))

# Compare means between combined conditions in wildtype only
ctrl <- means[(means$Sample == "control" & means$Stock == "wt"), "Mean"]
caut <- means[(means$Sample == "cauterized" & means$Stock == "wt"), "Mean"]

# Control vs Cauterized
contrast <- "ctrl vs caut"
test <- wilcox.test(caut, ctrl, alternative = "less", paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
mean_ctrl <- mean(ctrl)
mean_caut <- mean(caut)
delta <- mean_ctrl - mean_caut

test_row <- c(contrast = contrast,
              data = "means",
              sample = "wildtype",
              mean_ctrl = mean_ctrl,
              mean_caut = mean_caut,
              delta = sprintf("%.1f", delta),
              significant = pvalue < 0.05,
              pvalue = sprintf("%.9f", pvalue),
              statistic = sprintf("%s=%s", names(statistic), statistic),
              paired = FALSE)
tests <- rbind(tests, as.data.frame(t(test_row)))

write.csv(tests, file = "wt-tests.csv", row.names = FALSE)
```

