---
title: "Cephalic furrow depth in wildtype cauterizations"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: April 06, 2021
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Paper/1-analyses/germband-cauterizations")
library(ggplot2)
library(ggbeeswarm)
```

## Load measurements

```{r load}
# Load raw ROI measurements from ImageJ
results <- read.csv("1-wt/Lengths.csv")

# Discard cauterized embryo where the GB wasn't blocked from extending
results <- results[results$Embryo != "2021-03-24_E5",]
```

## Print summary

```{r summary}
print_stats <- function(dataframe, stock, sample) {
  filter_df <- (dataframe$Stock == stock & dataframe$Sample == sample)
  df <- dataframe[filter_df,]
  df <- droplevels(df)

  # Define row
  row <- c(stock,
           sample,
           mean(df$Length),
           sd(df$Length),
           length(df$Length),
           length(unique(df$Embryo))
           )
  
  # Summarize data
  print(sprintf("%s %s / area=%.3f±%.3f x 10^3 / n_folds=%i / n_embryos=%i",
                stock, sample,
                mean(df$Length),
                sd(df$Length),
                length(df$Length),
                length(unique(df$Embryo))
                ))
  return(row)
}

# Save stats to data frame
stats <- data.frame()

# Bind row by row
stats <- rbind(stats, print_stats(results, "wt", "control"))
stats <- rbind(stats, print_stats(results, "wt", "cauterized"))
stats <- rbind(stats, print_stats(results, "btd", "control"))
stats <- rbind(stats, print_stats(results, "btd", "cauterized"))
stats <- rbind(stats, print_stats(results, "eve", "control"))
stats <- rbind(stats, print_stats(results, "eve", "cauterized"))

# Add column names
colnames(stats) <- c("stock", "sample", "mean", "sd", "n_folds", "n_embryos")

# Convert columns to numeric
stats[, 3:5] <- apply(stats[, 3:5], 2, as.numeric)

# Round up
stats[, 3:4] <- round(stats[, 3:4], digits=1)

# Write table to file
write.table(format(stats, digits=2), file = "wt-stats.txt", quote = FALSE, row.names = FALSE, sep = "\t")
print(stats)
```

## Maximum depth of individual folds

```{r depth-ind}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(results, aes(factor(Sample, level = c("control", "cauterized")), Length, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  labs(x = "sample", y = "depth (µm)") +
  facet_wrap(~Stock) +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(size = 45, face = "italic"),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-length.png", width = 7, height = 4)
ggsave("4-plots/wt-length.svg", width = 7, height = 4)
```


## Mean furrow length by embryo

To attempt to normalize the furrow length per embryo, since their length does not
seem to be independent of each other, I'll take the length mean per embryo.

```{r mean-length}
# Mean of CF length per embryo
embryo_cf_mean <- aggregate(results$Length, list(results$Embryo), mean)
colnames(embryo_cf_mean) <- c("Embryo", "Mean")

# Standard deviation of CF length per embryo
embryo_cf_sd <- aggregate(results$Length, list(results$Embryo), sd)
colnames(embryo_cf_sd) <- c("Embryo", "SD")

# Get sample information
samples <- unique(results[, c("Stock", "Embryo", "Sample")])

# Merge datasets
means <- merge(samples, embryo_cf_mean, by = "Embryo", all.x = TRUE)
means <- merge(means, embryo_cf_sd, by = "Embryo", all.x = TRUE)

# Summary by sample
by(means[means$Sample == "control", c("Mean", "SD")], means[means$Sample == "control",]$Stock, summary)
print("------------------------------------------------------------------------------------")
by(means[means$Sample == "cauterized", c("Mean", "SD")], means[means$Sample == "cauterized",]$Stock, summary)
```

## Maximum depth of mean length by embryo

Due to the low n, we pooled the different genotypes together. There's a small
difference in length of about 15% (8 microns). But this needs more data to be
conclusive.

```{r depth-mean}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(means, aes(factor(Sample, level = c("control", "cauterized")), Mean, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  #geom_pointrange(aes(ymin=Mean-SD, ymax=Mean+SD), alpha = 0.8, size = 0.5, show.legend = FALSE, position = position_quasirandom(dodge.width = 0.75)) +
  #facet_wrap(~Stock) +
  labs(x = "condition", y = "depth (µm)") +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 25),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-mean-length.png", width = 4, height = 4)
ggsave("4-plots/wt-mean-length.svg", width = 4, height = 4)
```

```{r depth-mean-facet}
# Plot mean length of cephalic furrow per embryo (right/left sides)
ggplot(means[means$Stock != "eve",], aes(factor(Sample, level = c("control", "cauterized")), Mean, fill = Sample)) +
  geom_boxplot(aes(color = Sample), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  #geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  geom_pointrange(aes(ymin=Mean-SD, ymax=Mean+SD), alpha = 0.8, size = 0.5, show.legend = FALSE, position = position_quasirandom(width = 0.1)) +
  facet_wrap(~Stock) +
  labs(x = "condition", y = "depth (µm)") +
  scale_fill_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  scale_color_manual(values=c("#e69f03", "#56b4e9", "#cc79a7")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(size = 45, face = "italic"),
        panel.background = element_rect(colour = "black", size=1))
ggsave("4-plots/wt-mean-length-facet.png", width = 4, height = 4)
ggsave("4-plots/wt-mean-length-facet.svg", width = 4, height = 4)
```


## Statistics

```{r stats}
# Statistical testing
diff_micron <- mean(means[means$Sample == "control",]$Mean) - mean(means[means$Sample == "cauterized",]$Mean)
diff_pct <- diff_micron / mean(means[means$Sample == "cauterized",]$Mean) * 100

# Test if distribution is normal
shapiro.test(results[results$Sample == "control",]$Length)
shapiro.test(results[results$Sample == "cauterized",]$Length)
shapiro.test(means[means$Sample == "control",]$Mean)
shapiro.test(means[means$Sample == "cauterized",]$Mean)

# Test if means are different
t.test(results[results$Sample == "control",]$Length,
       results[results$Sample == "cauterized",]$Length,
       var.equal = FALSE)
t.test(means[means$Sample == "control",]$Mean,
       means[means$Sample == "cauterized",]$Mean,
       var.equal = FALSE)

# Test if means are different only in wt
t.test(results[(results$Sample == "control" & results$Stock == "wt"),]$Length,
       results[(results$Sample == "cauterized" & results$Stock == "wt"),]$Length,
       var.equal = FALSE)
t.test(means[(means$Sample == "control" & means$Stock == "wt"),]$Mean,
       means[(means$Sample == "cauterized" & means$Stock == "wt"),]$Mean,
       var.equal = FALSE)

# Test if control is greater than cauterized
wilcox.test(results[results$Sample == "control",]$Length,
            results[results$Sample == "cauterized",]$Length,
            alternative = "g")
wilcox.test(means[means$Sample == "control",]$Mean,
            means[means$Sample == "cauterized",]$Mean,
            alternative = "g")

# Test if control and cauterized have the same variance
var.test(results[results$Sample == "control",]$Length,
         results[results$Sample == "cauterized",]$Length)
var.test(means[means$Sample == "control",]$Mean,
         means[means$Sample == "cauterized",]$Mean)

```





