---
title: "Germ band speed estimation"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: September 12, 2024
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/germband-speed")
library(ggplot2)
library(ggbeeswarm)
```

## Import measurements

To estimate the germ band speed, we annotated the position of the tip of the germ band in two timepoints ten frames apart. We saved the coordinates as a RoiSet. Here is a sample file:

```
 	Label	X	Y	Slice
1	3D_btd-gap2_z3_t55s_E7.tif:0047-0113-1382:46	433	33	37
2	3D_btd-gap2_z3_t55s_E7.tif:0047-0113-1382:46	355	32	47
```

The distance traveled in X (microns) during ten frames gives the approximate speed of extension.

```{r import-measurements}
# List all txt files in ROIs folder
files = list.files(path = "1-rois", pattern="*.tsv", full.names = TRUE)

# Function for pre-processing
process <- function(tabfile) {
  
  # Read txt file
  df <- read.delim(tabfile)
  
  # Get label and dataset
  label <- df$Label[1]
  dataset <- strsplit(df$Label[1], ":")[[1]][1]
  
  # Clean up embryo name
  embryo <- gsub("3D_", "", dataset)
  embryo <- gsub(".tif", "", embryo)
  
  # Get X positions
  x_start <- df$X[1]
  x_end <- df$X[2]
  
  # Get sizes
  microns <- x_start - x_end
  
  # Get frames
  frame_start <- df$Slice[1]
  frame_end <- df$Slice[2]
  
  # Get frames
  frame_delta <- frame_end - frame_start
  
  # Get per frame distance
  microns_per_frame <- microns / frame_delta
  
  # Save as vector
  mv <- c(label = label,
          dataset = dataset,
          embryo = embryo,
          x_start = x_start,
          x_end = x_end,
          frame_start = frame_start,
          frame_end = frame_end,
          microns = microns,
          frame_delta = frame_delta,
          microns_per_frame = microns_per_frame
          )
  
  return(mv)
}

# Define empty data frame
gb_speed <- data.frame()

# Loop over files binding rows to data frame
for (tabfile in files) {
  gb_row <- process(tabfile)
  gb_speed <- rbind(gb_speed, as.data.frame(t(gb_row)))
}

# Save data frame to file
write.csv(gb_speed, file = "gb_speed.csv", row.names = FALSE)
```

## Merge datasets

```{r merge-datasets}

# Import main dataset
datasets <- read.csv("../../0-data/datasets.csv")

# Import germband-extension dataset
gb_extension <- read.csv("../germband-extension/gb_extension.csv")
germband <- read.csv("../germband-extension/germband.csv")
gb_extension <- cbind(gb_extension, germband[, c("time_tf_buck", "time_tf_deep")])

# Import mitotic-expansion dataset
md_expansion <- read.csv("../mitotic-expansion/md_expansion.csv")
mitotic <- read.csv("../mitotic-expansion/mitotic.csv")
md_expansion <- cbind(md_expansion, mitotic[, c("time_md_start", "gb_pct_md_start")])

# Merge landmarks with datasets
speed <- merge(gb_speed, gb_extension, by = "embryo", all.x = TRUE)
speed <- merge(speed, md_expansion, by = "embryo", all.x = TRUE)
speed <- merge(speed, datasets, by = "embryo", all.x = TRUE)

# Properly convert column types before any calculations
speed <- as.data.frame(lapply(speed, type.convert, as.is = FALSE))

```


## Calculate time

Now that the datasets are merged, we need to convert the frame information to time relative to gastrulation.

```{r calculate-time}

# Create columns with times (s and min)
speed$microns_per_second <- speed$microns / (speed$frame_delta * speed$timestep)
speed$microns_per_minute <- speed$microns / ((speed$frame_delta * speed$timestep) / 60)

# Save data frame to file
write.csv(speed, file = "speed.csv", row.names = FALSE)
```

## Create long format data frame

Convert the main data frame into the long format for plotting.

```{r reshape-long-format}

# Define variables for plotting
splong <- speed[, c("embryo",
                    "stock",
                    "zygosity",
                    "microns_per_minute",
                    "gb_pct_tf_buck",
                    "gb_pct_md_start",
                    "time_tf_buck",
                    "time_md_start")]

# Transform data frame to order the stock for plotting
splong$stock = factor(splong$stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
splong$zygosity = factor(splong$zygosity, levels = c("wildtype", "heterozygote", "homozygote", "double heterozygote", "btd homozygote", "stg homozygote", "double homozygote"))
splong <- splong[order(splong$stock),]
rownames(splong) <- 1:nrow(splong)

# Save long format data frame to file
write.csv(splong, file = "splong.csv", row.names = FALSE)
```

## Calculate mean and standard deviation

Create data frame with descriptive statistics for all the different samples and variables.

```{r values}

# Automatically aggregate data by columns
stats <- aggregate(microns_per_minute ~ stock + zygosity, splong, function(x)
  c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
  ))

# Makes column names nicer
stats <- do.call(data.frame, stats)

# Order rows by something meaningful (and re-index)
stats$stock = factor(stats$stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
stats <- stats[order(stats$stock),]
rownames(stats) <- 1:nrow(stats)

# Print and save data frame
stats
write.csv(stats, file = "stats.csv", row.names = FALSE)
```

## Exploratory plots

### Speed vs stock

```{r plot-gb-speed-all}
ggplot(splong, aes(stock, microns_per_minute, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  labs(x = "stocks", y = "speed (µm/min)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_speed_all.png", width = 20, height = 4)
ggsave("2-plots/gb_speed_all.svg", width = 20, height = 4)
```

### Speed vs variable

```{r plot-gb-speed-variables}

# time_tf_buck
ggplot(splong, aes(time_tf_buck, microns_per_minute, color = zygosity)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  labs(x = "time of buckling", y = "speed (µm/min)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_speed_vs_time_tf_buck.png", width = 9, height = 6)
ggsave("2-plots/gb_speed_vs_time_tf_buck.svg", width = 9, height = 6)

# time_md_start
ggplot(splong, aes(time_md_start, microns_per_minute, color = zygosity)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  labs(x = "time of mitosis", y = "speed (µm/min)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_speed_vs_time_md_start.png", width = 9, height = 6)
ggsave("2-plots/gb_speed_vs_time_md_start.svg", width = 9, height = 6)

# gb_pct_tf_buck
ggplot(splong, aes(gb_pct_tf_buck, microns_per_minute, color = zygosity)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  labs(x = "gb_pct at buckling", y = "speed (µm/min)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_speed_vs_gb_pct_tf_buck.png", width = 9, height = 6)
ggsave("2-plots/gb_speed_vs_gb_pct_tf_buck.svg", width = 9, height = 6)

# gb_pct_md_start
ggplot(splong, aes(gb_pct_md_start, microns_per_minute, color = zygosity)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03", "#0072b2", "#009966", "#f0e442", "#d55e00")) +
  labs(x = "gb_pct at mitosis", y = "speed (µm/min)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_speed_vs_gb_pct_md_start.png", width = 9, height = 6)
ggsave("2-plots/gb_speed_vs_gb_pct_md_start.svg", width = 9, height = 6)

```

## Main plot

```{r plot-gb-speed}
ggplot(splong[splong$stock %in% c("wt", "btd", "eve", "prd"),], aes(stock, microns_per_minute, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "speed (µm/min)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_speed_by_stock.png", width = 12, height = 4)
ggsave("2-plots/gb_speed_by_stock.svg", width = 12, height = 4)
```

# Statistics

```{r stats}
tests <- data.frame()

for (stock in c("btd", "eve", "prd", "stg")) {
  
  wt <- splong[splong$stock == "wt", ]$microns_per_minute
  het <- splong[(splong$stock == stock & splong$zygosity == "heterozygote"), ]$microns_per_minute
  hom <- splong[(splong$stock == stock & splong$zygosity == "homozygote"), ]$microns_per_minute
  
  # Heterozygotes vs homozygotes
  contrast = "het vs hom"
  test = wilcox.test(het, hom, paired = FALSE)
  pvalue = test$p.value
  statistic = test$statistic
  delta = mean(het) - mean(hom)
  
  test_row = c(
    contrast = contrast,
    stock = stock,
    value = "µm/min",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
  
  # Wildtype vs heterozygotes
  contrast = "wt vs het"
  test = wilcox.test(wt, het, paired = FALSE)
  pvalue = test$p.value
  statistic = test$statistic
  delta = mean(wt) - mean(het)
  
  test_row = c(
    contrast = contrast,
    stock = stock,
    value = "µm/min",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
  
  # Wildtype vs homozygotes
  contrast = "wt vs hom"
  test = wilcox.test(wt, hom, paired = FALSE)
  pvalue = test$p.value
  statistic = test$statistic
  delta = mean(wt) - mean(hom)
  
  test_row = c(
    contrast = contrast,
    stock = stock,
    value = "µm/min",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
}

tests
write.csv(tests, file = "tests.csv", row.names = FALSE)
```
