---
title: "Tissue dynamics after laser cuts"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: September 27, 2024
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/tissue-cuts")
library(ggplot2)
library(ggbeeswarm)
```

## Import measurements

To analyze the tissue dynamics after laser cuts, we tracked a pair of vertices of non-ablated cells sitting at the edges of the cut every 50 frames. Then, we measured the euclidean distance between the vertices over time.

Here is a sample data file from TrackMate:

```         
LABEL	ID	TRACK_ID	QUALITY	POSITION_X	POSITION_Y	POSITION_Z	POSITION_T	FRAME	RADIUS	VISIBILITY
Label	Spot ID	Track ID	Quality	X	Y	Z	T	Frame	Radius	Visibility
Label	Spot ID	Track ID	Quality	X	Y	Z	T	Frame	R	Visibility
			(quality)	(micron)	(micron)	(micron)	(sec)		(micron)	
ID2853067	2853067	3	-1	51.7865123753519	72.5230220892749	0	0	0	5	1
ID2853108	2853108	1	-1	63.1769895421547	62.1547672323134	0	0	0	5	1
ID2853069	2853069	3	-1	52.3706394095469	74.1293714333112	0	25.4814602	49	5	1
ID2853110	2853110	1	-1	64.783338886191	63.6150848178009	0	25.4814602	49	5	1
```

Import the data and generate a tracking data frame.

```{r import-measurements}

# List all txt files in ROIs folder
files <- list.files(path = "1-tracking", pattern="*.csv", full.names = TRUE)

# Define empty data frame
tracking <- data.frame()

# Loop over files binding rows to data frame
for (tabfile in files) {
  
  # Read CSV file
  df <- read.csv(tabfile)
  
  # Skip first three rows (labels)
  df <- df[4:nrow(df),]
  
  # Define dataset column
  df$DATASET <- strsplit(strsplit(tabfile, "/")[[1]][2], "_")[[1]][1]
  
  # Add to data frame
  tracking <- rbind(tracking, df)
}

# Properly convert column types before any calculations
tracking <- as.data.frame(lapply(tracking, type.convert, as.is = FALSE))

# Define condition column
tracking$DATASET <- as.character(tracking$DATASET)
tracking$CONDITION <- ifelse(startsWith(tracking$DATASET, "c"), "control", "ablated")

# Define sample column
tracking$SAMPLE <- NA

# Define standardized sample names
for (condition in unique(tracking$CONDITION)) {
  
  # Get subset with condition and sort properly
  datasets <- unique(tracking[tracking$CONDITION == condition, ]$DATASET)
  datasets <- datasets[order(as.numeric(gsub("[^0-9]", "", datasets)))]
  
  # Initialize counter
  counter = 1
  
  # Loop over datasets
  for (dataset in datasets) {
    
    # Define sample name
    sample <- sprintf("%s%s", condition, counter)
    
    # Set sample name
    tracking[tracking$DATASET == dataset, ]$SAMPLE <- sample
    
    # Increment counter
    counter <- counter + 1
    
  }
}

# Set factors for plotting
tracking$TRACK_ID <- factor(tracking$TRACK_ID, levels = c("0", "1"))
tracking$DATASET <- factor(tracking$DATASET, levels = c("cONTROL1", "cONTROL3", "cONTROL4", "Lateral2", "Lateral4", "Lateral10S"))
tracking$SAMPLE <- factor(tracking$SAMPLE, levels = c("control1", "control2", "control3", "ablated1", "ablated2", "ablated3"))
tracking$CONDITION <- factor(tracking$CONDITION, levels = c("control", "ablated"))
tracking <- tracking[order(tracking$DATASET, tracking$TRACK_ID, tracking$POSITION_T),]
rownames(tracking) <- 1:nrow(tracking)

# Save data frame to file
write.csv(tracking, file = "tracking.csv", row.names = FALSE)

```

## Calculate absolute and relative distances

Using the XY coordinates, calculate the euclidean distance between the two points. And generate a new data frame with the distances for plotting.

```{r calculate-distances}

# Calculate the euclidean distance between two points
get_euclidean <- function(ax, ay, bx, by) {
  sqrt((ax - bx)^2 + (ay - by)^2)
}

# Define distances data frame
distances <- data.frame()

# Build distances by dataset
for (dataset in unique(tracking$DATASET)) {
  
  # Subset dataset
  df <- tracking[tracking$DATASET == dataset, ]
  
  # Define condition and sample
  condition <- as.character(unique(df$CONDITION))
  sample <- as.character(unique(df$SAMPLE))
  
  # Iterate per frame
  for (frame in unique(df$FRAME)) {
    
    # Get points 1 and 2
    p1 <- df[df$FRAME == frame & df$TRACK_ID == "0",]
    p2 <- df[df$FRAME == frame & df$TRACK_ID == "1",]
    
    # Calculate distance between 1 and 2
    dist_p1p2 <- get_euclidean(p1$POSITION_X, p1$POSITION_Y, p2$POSITION_X, p2$POSITION_Y)
    
    # Define initial distance at frame 0
    if (frame == 0) {
      init_p1p2 <- dist_p1p2
    }
    
    # Calculate percentage between 1 and 2
    pct_p1p2 <- (dist_p1p2 / init_p1p2) * 100
    
    # Define new data frame row
    row <- c(
      dataset = dataset,
      sample = sample,
      condition = condition,
      frame = frame,
      time_s = p1$POSITION_T,
      time_min = p1$POSITION_T / 60,
      distance = dist_p1p2,
      percentage = pct_p1p2
    )
    
    # Bind to data frame
    distances <- rbind(distances, as.data.frame(t(row)))
  }
}

# Properly convert column types before any calculations
distances <- as.data.frame(lapply(distances, type.convert, as.is = FALSE))

# Set factors for plotting
distances$dataset <- factor(distances$dataset, levels = c("cONTROL1", "cONTROL3", "cONTROL4", "Lateral2", "Lateral4", "Lateral10S"))
distances$sample <- factor(distances$sample, levels = c("control1", "control2", "control3", "ablated1", "ablated2", "ablated3"))
distances$condition <- factor(distances$condition, levels = c("control", "ablated"))
distances <- distances[order(distances$dataset, distances$frame),]
rownames(distances) <- 1:nrow(distances)

# Save data frame to file
write.csv(distances, file = "distances.csv", row.names = FALSE)

```


## Exploratory plots

### Distance between vertices over time by dataset with facets

```{r distance-vs-time-by-dataset-facets}

# Ablations:
# Lateral2 frame 8
# Lateral4 frame 57
# Lateral10S frame 9

ggplot(distances, aes(x = time_min, y = distance, group = sample, color = sample)) +
  geom_rect(data = data.frame(xmin = 8*0.52/60, xmax = 57*0.52/60, ymin = -1, ymax = 22, condition = factor("ablated", levels = c("control", "ablated"))), aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), alpha = 0.5, fill = "#cccccc", inherit.aes = FALSE) +
  geom_line(linewidth = 1, show.legend = FALSE) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  facet_grid(cols = vars(condition), scales = "free_x") +
  labs(x = "time (min)", y = "distance (µm)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 25),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))

ggsave("2-plots/distance_vs_time_by_dataset_facets.png", width = 14, height = 7)
ggsave("2-plots/distance_vs_time_by_dataset_facets.svg", width = 14, height = 7)

```

### Relative distance between vertices over time by dataset with facets

```{r delta-vs-time-by-dataset-facets}

ggplot(distances, aes(x = time_min, y = percentage, group = sample, color = sample)) +
  geom_rect(data = data.frame(xmin = 8*0.52/60, xmax = 57*0.52/60, ymin = -1, ymax = 125, condition = factor("ablated", levels = c("control", "ablated"))), aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), alpha = 0.5, fill = "#cccccc", inherit.aes = FALSE) +
  geom_line(linewidth = 1, show.legend = FALSE) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE) +
  facet_grid(cols = vars(condition), scales = "free_x") +
  labs(x = "time (min)", y = "relative distance (%)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 25),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))

ggsave("2-plots/delta_vs_time_by_dataset_facets.png", width = 10, height = 6)
ggsave("2-plots/delta_vs_time_by_dataset_facets.svg", width = 10, height = 6)
```
### Relative distance between vertices over time by condition with facets

```{r delta-vs-time-by-condition-smooth-facets}
ggplot(distances, aes(x = time_min, y = percentage, group = sample, color = condition, fill = condition)) +
  geom_line(alpha = 0.2, linewidth = 0.5, show.legend = FALSE) +
  geom_point(alpha = 0.2, size = 3, show.legend = FALSE) +
  geom_smooth(aes(group = condition), linewidth = 2, alpha = 0.3, show.legend = FALSE) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_y_continuous(limits = c(0,130)) +
  facet_grid(cols = vars(condition), scales = "free_x") +
  labs(x = "time (min)", y = "relative distance (%)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 45),
        title = element_text(size = 25),
        legend.text = element_text(size = 25),
        axis.text = element_text(size = 25),
        strip.background = element_blank(),
        strip.text = element_blank(),
        panel.background = element_rect(colour = "black", size=1))

ggsave("2-plots/delta_vs_time_by_condition_smooth_facets.png", width = 8, height = 4)
ggsave("2-plots/delta_vs_time_by_condition_smooth_facets.svg", width = 8, height = 4)
```

### Relative distance between vertices over time by condition

```{r delta-vs-time-by-condition-smooth}
ggplot(distances, aes(x = time_min, y = percentage, group = sample, color = condition, fill = condition)) +
  geom_line(alpha = 0.2, linewidth = 0.5, show.legend = FALSE) +
  geom_point(alpha = 0.2, size = 2, show.legend = TRUE) +
  geom_smooth(aes(group = condition), linewidth = 1.5, alpha = 0.2, show.legend = FALSE) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_y_continuous(limits = c(0,130)) +
  labs(x = "time (min)", y = "relative distance (%)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 25),
        legend.text = element_text(size = 25),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))

ggsave("2-plots/delta_vs_time_by_condition_smooth.png", width = 10, height = 6)
ggsave("2-plots/delta_vs_time_by_condition_smooth.svg", width = 10, height = 6)
```

