---
title: "Maximum depth of folds in cephalic furrow mutants"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: May 31, 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/fold-depth")
library(ggplot2)
library(ggbeeswarm)
```

## Import measurements

```{r import-measurements}
# Load sample info and merge with results
samples <- read.delim("samples.csv", sep = ",")

# Load raw ROI measurements from ImageJ
results <- read.delim("results.txt")

# Remove column with row number and angle
results <- subset(results, select = -X)
results <- subset(results, select = -Angle)

# Replace label with proper label
results$Label <- paste0(sapply(strsplit(results$Label, ":"), "[", 1),
                        "_", sapply(strsplit(results$Label, ":"), "[", 2))

# Merge samples with results
results <- merge(samples, results, by = "Label", all.x = TRUE)

# Add column for embryo sorting
results$E <- as.numeric(sub(".*_E(\\d+)$", "\\1", results$Embryo))

# Transform data frame to order the stock for plotting
results$Stock <- factor(results$Stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
results$Genotype <- factor(results$Genotype, levels = c("wildtype", "heterozygote", "homozygote", "double heterozygote", "btd homozygote", "stg homozygote", "double homozygote"))
results$Position <- factor(results$Position, levels = c("ant", "mid", "post"))

results <- results[order(results$Stock, results$Genotype, results$Position, results$E),]
rownames(results) <- 1:nrow(results)

# Save data frame to file
write.csv(results, file = "combined_results.csv", row.names = FALSE)
```

## Create main fold-depth dataset

The dataset only include ROIs scored as true folds (after revision) and remove any NA values without area measurements.

```{r create-depth-results}
# For downstream area analysis, only consider TRUE folds without NA
depth <- results[results$Fold == TRUE & !is.na(results$Length), ]

# Save data frame to file
write.csv(depth, file = "depth.csv", row.names = FALSE)
```

## Create total fold depth per embryo

```{r calculate-total-depth-per-embryo}

# Total folds and folded area per embryo
depth_total <- aggregate(
  Length ~ Embryo + Stock + Genotype + Type,
  depth,
  function(x) c(
    count = length(x),
    sum = sum(x),
    mean = mean(x),
    sd = sd(x)
  ))

# Makes column names nicer
depth_total <- do.call(data.frame, depth_total)
depth_total <- depth_total[order(depth_total$Stock, depth_total$Genotype, depth_total$Type),]

# Save data frame to file
write.csv(depth_total, file = "depth_total.csv", row.names = FALSE)

depth_total
```

## Summarize total depth per embryo by genotype

```{r summarize-total-depth-per-embryo-by-genotype}

# Total folds and folded area per Stock and Genotype
stats_total <- aggregate(
  cbind(Length.sum, Length.mean) ~ Stock + Genotype,
  depth_total,
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
    ))

# Makes column names nicer
stats_total <- do.call(data.frame, stats_total)
stats_total <- subset(stats_total, select = -Length.mean.count)
colnames(stats_total)[colnames(stats_total) == "Length.sum.count"] <- "Embryos"
stats_total <- stats_total[order(stats_total$Stock, stats_total$Genotype),]

# Save data frame to file
write.csv(stats_total, file = "stats_total.csv", row.names = FALSE)

stats_total
```

## Summarize individual fold depth per type

```{r summarize-individual-fold-depth-per-type}

# Total folds and folded area per Stock and Genotype
stats_type <- aggregate(
  Length ~ Stock + Genotype + Type,
  depth,
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
    ))

# Makes column names nicer
stats_type <- do.call(data.frame, stats_type)
stats_type <- stats_type[order(stats_type$Stock, stats_type$Genotype, stats_type$Type),]

# Save data frame to file
write.csv(stats_type, file = "stats_type.csv", row.names = FALSE)

stats_type
```

## Summarize individual fold depth per type by position

```{r summarize-individual-fold-depth-per-type-by-position}

# Total folds and folded area per Stock and Genotype
stats_position <- aggregate(
  Length ~ Stock + Genotype + Type + Position,
  depth,
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
    ))

# Makes column names nicer
stats_position <- do.call(data.frame, stats_position)
stats_position <- stats_position[order(stats_position$Stock, stats_position$Genotype, stats_position$Position),]

# Save data frame to file
write.csv(stats_position, file = "stats_position.csv", row.names = FALSE)

stats_position
```

## Plot mean total fold depth by stocks

```{r plot-mean-total-depth}
# Plot stocks separately
ggplot(depth_total, aes(Genotype, Length.mean, fill = Genotype, shape = Type)) +
  geom_boxplot(aes(color = Genotype), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 5, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  labs(x = "mutants", y = "depth") +
  facet_wrap(~Stock, strip.position = "top") +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 40),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(face = "italic"),
        panel.background = element_rect(colour = "black", linewidth = 1))
ggsave("3-plots/total_depth_facet.png", width = 12, height = 6)
ggsave("3-plots/total_depth_facet.svg", width = 12, height = 6)

```

## Plot individual fold depth by stocks

```{r plot-individual-depth}
# Plot stocks separately
ggplot(depth, aes(Genotype, Length, fill = Genotype, shape = Type)) +
  geom_boxplot(aes(color = Genotype), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 5, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 2.5)) +
  labs(x = "mutants", y = "depth") +
  facet_wrap(~Stock, strip.position = "top") +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 40),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(face = "italic"),
        panel.background = element_rect(colour = "black", linewidth = 1))
ggsave("3-plots/depth_facet.png", width = 12, height = 6)
ggsave("3-plots/depth_facet.svg", width = 12, height = 6)

```

## Statistics

```{r stats}
tests <- data.frame()

for (stock in c("btd", "eve")) {
  
  # Total mean
  het <- depth_total[(depth_total$Stock == stock & depth_total$Genotype == "heterozygote" & depth_total$Type == "CF"), "Length.mean"]
  hom <- depth_total[(depth_total$Stock == stock & depth_total$Genotype == "homozygote" & depth_total$Type == "EF"), "Length.mean"]
  
  # Heterozygotes vs Homozygotes
  contrast <- "het vs hom"
  test <- wilcox.test(het, hom, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(het) - mean(hom)
  
  test_row <- c(
    contrast = contrast,
    subject = stock,
    value = "total",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
  
  # Individual folds
  het <- depth[(depth$Stock == stock & depth$Genotype == "heterozygote" & depth$Type == "CF"), "Length"]
  hom <- depth[(depth$Stock == stock & depth$Genotype == "homozygote" & depth$Type == "EF"), "Length"]
  
  # Heterozygotes vs Homozygotes
  contrast <- "het vs hom"
  test <- wilcox.test(het, hom, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(het) - mean(hom)
  
  test_row <- c(
    contrast = contrast,
    subject = stock,
    value = "individual",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))

}

for (genotype in c("heterozygote", "homozygote")) {
  
  if (genotype == "heterozygote") {
    type <- "CF"
  } else if (genotype == "homozygote") {
    type <- "EF"
  }
    
  # Total mean
  btd <- depth_total[(depth_total$Stock == "btd" & depth_total$Genotype == genotype & depth_total$Type == type), "Length.mean"]
  eve <- depth_total[(depth_total$Stock == "eve" & depth_total$Genotype == genotype & depth_total$Type == type), "Length.mean"]
  
  # Heterozygotes vs Homozygotes
  contrast <- "btd vs eve"
  test <- wilcox.test(btd, eve, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(btd) - mean(eve)
  
  test_row <- c(
    contrast = contrast,
    subject = sprintf("%s %s", genotype, type),
    value = "total",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
  
  # Individual folds
  btd <- depth[(depth$Stock == "btd" & depth$Genotype == genotype & depth$Type == type), "Length"]
  eve <- depth[(depth$Stock == "eve" & depth$Genotype == genotype & depth$Type == type), "Length"]
  
  # Heterozygotes vs Homozygotes
  contrast <- "btd vs eve"
  test <- wilcox.test(btd, eve, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(btd) - mean(eve)
  
  test_row <- c(
    contrast = contrast,
    subject = sprintf("%s %s", genotype, type),
    value = "individual",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))

}

# Save test data frame to file
write.csv(tests, file = "tests.csv", row.names = FALSE)
tests
```
