---
title: "Folded area during ectopic folding"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: February 18, 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/folded-area")
library(ggplot2)
library(ggbeeswarm)
```

## Import measurements

We exported the area measurements from ROIs on cartographic projections and corrected the values to microns.

Sample of `results.txt`:

```
 	Label	Area	StdDev	X	Y	XM	YM	Perim.	BX	BY	Width	Height	Major	Minor	Angle	Circ.	Feret	IntDen	Median	Skew	Kurt	RawIntDen	FeretX	FeretY	FeretAngle	MinFeret	AR	Round	Solidity
1	btd-gap2_z3_t55s_E12_CARE_s24_0x0_w1724_h1106_fliph_f48-117.1.roi:1	98543.000	217.480	499.839	515.725	500.504	520.260	1823.388	395.000	180.000	200.000	744.000	712.253	176.158	87.248	0.372	746.496	120829781.000	1185	0.972	1.586	120829781.000	477	924	85.313	198.872	4.043	0.247	0.893
2	btd-gap2_z3_t55s_E7_CARE_s20_0x0_w1721_h1029_flipn_f23-92.1.roi:1	7095.000	103.782	253.389	841.867	252.628	839.782	459.002	213.000	746.000	85.000	171.000	170.569	52.962	110.872	0.423	177.775	7733816.000	1071	1.160	1.531	7733816.000	221	746	107.008	68.088	3.221	0.311	0.787
```

Sample of `corrected_results.txt`:

```
File "ROI Area"
rgap_1b_z3_t60s_E15_s20_0x70_w1777_h980_fliph_f45-114.1.roi 261.57975999999996
rgap_1b_z3_t60s_E15_s20_0x70_w1777_h980_fliph_f45-114.2.roi 10251.997106000003
```

```{r import-measurements}
# Load sample info, remove comments
samples <- read.delim("samples.csv", sep = ",")
samples <- subset(samples, select = -Comment)

# Load raw ROI measurements from ImageJ
results <- read.delim("results.txt")

# Remove column with row number
results <- subset(results, select = -X.1)

# Remove slice number from the labels
results$Label <- gsub(":1", "", results$Label)

# Merge samples with results
results <- merge(samples, results, by = "Label", all.x = TRUE)

# Load corrected area measurements for cartographic projections
corrected <- read.delim("corrected_results.txt", sep = "")
colnames(corrected) <- c("Label", "AreaCorr")

# Merge corrected area to results
results <- merge(results, corrected, by = "Label", all.x = TRUE)

# Add column for embryo sorting
results$E <- as.numeric(sub(".*_E(\\d+)$", "\\1", results$Embryo))

# Transform data frame to order the stock for plotting
results$Stock <- factor(results$Stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
results$Genotype <- factor(results$Genotype, levels = c("wildtype", "heterozygote", "homozygote", "double heterozygote", "btd homozygote", "stg homozygote", "double homozygote"))
results$Position <- factor(results$Position, levels = c("ant", "mid", "post"))

results <- results[order(results$Stock, results$Genotype, results$Position, results$E),]
rownames(results) <- 1:nrow(results)

# Save data frame to file
write.csv(results, file = "combined_results.csv", row.names = FALSE)
```

## Create main folded-area dataset

The dataset only include ROIs scored as true folds (after revision) and remove any NA values without area measurements.

```{r create-area-results}
# For downstream area analysis, discard FALSE folds and keep only folds with a measurement
area <- results[results$Discard == FALSE & !is.na(results$AreaCorr), ]

# Save data frame to file
write.csv(area, file = "area.csv", row.names = FALSE)
```

## Create total folded-area per embryo

```{r calculate-total-area-per-embryo}
# Total folds and folded area per embryo
area_total <- aggregate(AreaCorr ~ Embryo + Stock + Genotype,
                                   area, function(x) c(
                                     count = length(x),
                                     sum = sum(x)))

# Makes column names nicer
area_total <- do.call(data.frame, area_total)
colnames(area_total)[colnames(area_total) == "AreaCorr.count"] <- "Folds.count"

# Save data frame to file
write.csv(area_total, file = "area_total.csv", row.names = FALSE)
```

## Summarize total area per embryo by genotype

```{r summarize-total-area-per-embryo-by-genotype}

# Total folds and folded area per Stock and Genotype
stats_total <- aggregate(
  cbind(Folds.count, AreaCorr.sum) ~ Stock + Genotype,
  area_total,
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
    ))

# Makes column names nicer
stats_total <- do.call(data.frame, stats_total)
stats_total <- subset(stats_total, select = -AreaCorr.sum.count)
colnames(stats_total)[colnames(stats_total) == "Folds.count.count"] <- "Embryos"
stats_total <- stats_total[order(stats_total$Stock),]

# Save data frame to file
write.csv(stats_total, file = "stats_total.csv", row.names = FALSE)

stats_total
```

## Summarize individual fold area per type

```{r summarize-individual-fold-area-per-type}

# Total folds and folded area per Stock and Genotype
stats_type <- aggregate(
  AreaCorr ~ Stock + Genotype + Type,
  area,
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
    ))

# Makes column names nicer
stats_type <- do.call(data.frame, stats_type)
stats_type <- stats_type[order(stats_type$Stock),]

# Save data frame to file
write.csv(stats_type, file = "stats_type.csv", row.names = FALSE)

stats_type
```

## Summarize individual fold area per type by position

```{r summarize-individual-fold-area-per-type-by-position}

# Total folds and folded area per Stock and Genotype
stats_position <- aggregate(
  AreaCorr ~ Stock + Genotype + Type + Position,
  area,
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
    ))

# Makes column names nicer
stats_position <- do.call(data.frame, stats_position)
stats_position <- stats_position[order(stats_position$Stock),]

# Save data frame to file
write.csv(stats_position, file = "stats_position.csv", row.names = FALSE)

stats_position
```


## Total folded area per embryo

Plot the total area of folded tissue per embryo of different genotypes.

```{r total-area-per-embryo}
ggplot(area_total, aes(Stock, AreaCorr.sum, fill = Genotype)) +
  geom_boxplot(aes(color = Genotype), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_y_continuous(labels = function(x) x / 1000) +
  labs(x = "stocks", y = expression(paste("folded area per embryo ", (µm^2)~x~10^3))) +
  facet_grid(cols = vars(Stock), scales = "free_x") +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1),
        )
ggsave("4-plots/total-area-per-embryo.png", width = 12, height = 4)
ggsave("4-plots/total-area-per-embryo.svg", width = 12, height = 4)
```

## Area of individual ectopic folds

```{r individual-area-per-ectopic-fold}
ggplot(area[area$Type == "EF",], aes(Genotype, AreaCorr, fill = Genotype)) +
  geom_boxplot(aes(color = Genotype), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 5, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_y_continuous(labels = function(x) x / 1000) +
  labs(x = "stocks", y = expression(paste("area ", x~10^3~(µm^2)))) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 40),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("4-plots/individual-area-per-ectopic-fold.png", width = 7, height = 6)
ggsave("4-plots/individual-area-per-ectopic-fold.svg", width = 7, height = 6)
```


## Area of individual ectopic folds per position

```{r area-individual-folds-per-position}
ggplot(area, aes(Stock, AreaCorr, fill = Genotype)) +
  geom_boxplot(aes(color = Genotype), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 3)) +
  scale_y_continuous(labels = function(x) x / 1000) +
  labs(x = "stocks", y = expression(paste("area of individual folds ", (µm^2)~x~10^3))) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  facet_grid(cols = vars(Position), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 30),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1),
        axis.text.x = element_text(face = "italic"),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        )
ggsave("4-plots/individual-area-per-position.png", width = 12, height = 4)
ggsave("4-plots/individual-area-per-position.svg", width = 12, height = 4)
```

## Area of individual folds per type

```{r individual-area-per-type}
ggplot(area, aes(Type, AreaCorr, fill = Genotype)) +
  geom_boxplot(aes(color = Genotype), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_y_continuous(labels = function(x) x / 1000) +
  labs(x = "fold type", y = expression(paste("area ", x~10^3~(µm^2)))) +
  facet_grid(cols = vars(Stock), scales = "free_x") +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        strip.text.x = element_text(face = "italic"),
        panel.background = element_rect(colour = "black", linewidth = 1),
        )
ggsave("4-plots/individual-area-per-type.png", width = 13, height = 4)
ggsave("4-plots/individual-area-per-type.svg", width = 13, height = 4)
```

# Statistics

```{r stats}
tests <- data.frame()

for (stock in c("btd", "eve", "prd")) {
  
  wt <- area_total[area_total$Stock == "wt", "AreaCorr.sum"]
  het <- area_total[(area_total$Stock == stock & area_total$Genotype == "heterozygote"), "AreaCorr.sum"]
  hom <- area_total[(area_total$Stock == stock & area_total$Genotype == "homozygote"), "AreaCorr.sum"]
  
  # Heterozygotes vs Homozygotes
  contrast <- "het vs hom"
  test <- wilcox.test(het, hom, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(het) - mean(hom)
  
  test_row <- c(
    contrast = contrast,
    stock = stock,
    value = "total",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
  
  # Wildtype vs Heterozygotes
  contrast <- "wt vs het"
  test <- wilcox.test(wt, het, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(wt) - mean(het)
  
  test_row <- c(
    contrast = contrast,
    stock = stock,
    value = "total",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
  
  # Wildtype vs Homozygotes
  contrast <- "wt vs hom"
  test <- wilcox.test(wt, hom, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(wt) - mean(hom)
  
  test_row <- c(
    contrast = contrast,
    stock = stock,
    value = "total",
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))
}

# Compare all EFs by genotype below
wt <- area[area$Type == "EF" & area$Genotype == "wildtype", "AreaCorr"]
het <- area[area$Type == "EF" & area$Genotype == "heterozygote", "AreaCorr"]
hom <- area[area$Type == "EF" & area$Genotype == "homozygote", "AreaCorr"]

# Heterozygotes vs Homozygotes
contrast <- "het vs hom"
test <- wilcox.test(het, hom, paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
delta <- mean(het) - mean(hom)

test_row <- c(
  contrast = contrast,
  stock = "all",
  value = "EFs",
  delta = sprintf("%.1f", delta),
  significant = pvalue < 0.05,
  pvalue = sprintf("%.9f", pvalue),
  statistic = sprintf("%s=%s", names(statistic), statistic),
  paired = FALSE
)

tests <- rbind(tests, as.data.frame(t(test_row)))


# Wildtype vs Heterozygotes
contrast <- "wt vs het"
test <- wilcox.test(wt, het, paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
delta <- mean(wt) - mean(het)

test_row <- c(
  contrast = contrast,
  stock = "all",
  value = "EFs",
  delta = sprintf("%.1f", delta),
  significant = pvalue < 0.05,
  pvalue = sprintf("%.9f", pvalue),
  statistic = sprintf("%s=%s", names(statistic), statistic),
  paired = FALSE
)

tests <- rbind(tests, as.data.frame(t(test_row)))

# Wildtype vs Homozygotes
contrast <- "wt vs hom"
test <- wilcox.test(wt, hom, paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
delta <- mean(wt) - mean(hom)

test_row <- c(
  contrast = contrast,
  stock = "all",
  value = "EFs",
  delta = sprintf("%.1f", delta),
  significant = pvalue < 0.05,
  pvalue = sprintf("%.9f", pvalue),
  statistic = sprintf("%s=%s", names(statistic), statistic),
  paired = FALSE
)

tests <- rbind(tests, as.data.frame(t(test_row)))

# Save test data frame to file
write.csv(tests, file = "tests.csv", row.names = FALSE)
tests
```

