---
title: "Invaginated area during ectopic folding"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: February 18, 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Paper/1-analyses/folded-area")
# Load ggplot
library(ggplot2)
```

## Analysis

We think the ectopic folding in cephalic furrow mutants occur because the
epithelial tissue is under compression, and buckles once the forces caused by
the mitotic domains and germ band extension cross a certain threshold. We
notice there is some sort of compensation in the ectopic folding events---when
an anterior fold is deep, no other folds form; when the anterior fold is
shallow, other folds form. We thus expect the total invaginated area would be
similar between mutant embryos, despite the variability in the folding events.

To estimate the invaginated area we used the cartographic projections of btd
and eve embryos (see 0-data), both heterozygotes and homozygotes, and visually
annotate the blastoderm cells that are invaginated into ectopic folding events.
This is done by tracking the cells that delimit the borders of ectopic
folds. We use ImageJ/Fiji selection brush to delineate the area of invagination
and save it as a ROI (see 1-rois). Using a script we compute the measurements
for individual folds, correct for the cartographic projection area and export
the data as csv file.

## Directories

- `0-data`: single-slice, cropped imsane projections
- `1-rois`: imagej rois for individual folds visually annotated (>1 per embryo)
- `2-cylinder`: correction matrix for imsane projections
- `3-tracking`: in case we need actual cell tracking

Filenames contain all the information from the original projection. For
example: `eve-Gap1_t60s_z3_E4_CARE_s20_0x334_w1850_h1040_fliph_f15-84.tif`

- `s20` = slice 20 in Z
- `0x334` = XY origin for cropping
- `w1850` = width of crop window (=projection width)
- `h1040` = height of crop window
- `fliph` = crop was flipped horizontally (flipn means no flipping)
- `f15-84` = subset of frames for this analysis (=70 frames)

## Load measurements

```{r load}
# Load raw ROI measurements from ImageJ
results <- read.delim("results.txt")

# Remove slice number from the label
results$Label <- gsub(":1", "", results$Label)

# Split label to get embryo and sapply to get all first elements of the split...
results$Embryo <- sapply(strsplit(results$Label, "_CARE"), "[", 1)

# Add columns with sample information (needs manual update for new ROIs)
results$Stock <- substring(results$Label, 1, 3)

# Load sample info and merge with results
samples <- read.delim("samples.csv", sep = ",")
results <- merge(results, samples, by = "Label", all.x = TRUE)

# Load corrected area measurements for cartographic projections
corrected <- read.delim("corrected_results.txt", sep = "")
colnames(corrected) <- c("Label", "AreaCorr")

#corrected_area$Embryo <- sapply(strsplit(corrected_area$File, "_CARE"), "[", 1)
results <- merge(results, corrected, by = "Label", all.x = TRUE)

# Sum total folded area
area <- aggregate(results$Area, list(results$Embryo), sum)
colnames(area) <- c("Embryo", "Area")

# Sum total folded area corrected
areacorr <- aggregate(results$AreaCorr, list(results$Embryo), sum)
colnames(areacorr) <- c("Embryo", "AreaCorr")

# Count number of folds
nfold <- data.frame(table(results$Embryo))
colnames(nfold) <- c("Embryo", "Folds")

# Get sample information
embryos <- unique(results[, c("Embryo", "Stock", "Sample", "Genotype")])

# Merge datasets
total <- merge(area, areacorr, by = "Embryo", all.x = TRUE)
total <- merge(total, nfold, by = "Embryo", all.x = TRUE)
total <- merge(total, embryos, by = "Embryo", all.x = TRUE)

```

## Area plot

```{r area}
# Plot raw area of individual folds
ggplot(results, aes(factor(Sample, level = c("control", "btd", "eve")), Area, colour = Sample)) +
  geom_violin() + geom_jitter(alpha = 0.8, width = 0.15)

# Plot corrected area of individual folds
ggplot(results, aes(factor(Sample, level = c("control", "btd", "eve")), AreaCorr, colour = Sample)) +
  geom_violin() + geom_jitter(alpha = 0.8, width = 0.15) +
  labs(title = "Invaginated area of cephalic furrow compared to ectopic folds", x = "sample", y = expression(micrometers^2))
ggsave("folded-area_mutants.png")

# Plot corrected area of individual folds
ggplot(results, aes(Genotype, AreaCorr, fill = Genotype)) +
  geom_violin(draw_quantiles = c("0.5"), show.legend = FALSE) +
  geom_jitter(alpha = 0.8, width = 0.1, size = 2, show.legend = FALSE) +
  labs(x = "mutants", y = expression(paste("folded area ", (µm^2)))) +
  facet_wrap(~Stock, strip.position = "top") +
  scale_color_manual(values=c("#0072b2", "#e69f03")) +
  scale_fill_manual(values=c("#0072b2", "#e69f03")) +
  theme_classic() +
  theme(text =  element_text(size = 25),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_text(face = "italic"),
        panel.background = element_rect(colour = "black", size=1))
ggsave("folded-area_mutants_facet.png", width = 14, height = 5)


# Plot corrected area of individual folds by genotype
ggplot(results, aes(Genotype, AreaCorr, colour = Genotype)) + geom_violin(draw_quantiles = c("0.5")) + geom_jitter(alpha = 0.8, width = 0.15) +
  labs(x = "Genotype", y = expression(micrometers^2))
ggsave("folded-area_all.png", width = 5, height = 3)

# Plot raw area of folding per embryo
ggplot(total, aes(factor(Genotype, level = c("control", "btd", "eve")), Area, colour = Sample)) + geom_violin() + geom_jitter(alpha = 0.8, width = 0.15)

# Plot corrected area of folding per embryo
ggplot(total, aes(factor(Sample, level = c("control", "btd", "eve")), AreaCorr, colour = Sample)) +
  geom_violin() + geom_jitter(alpha = 0.8, width = 0.15) +
  labs(title = "Total area of folded tissue", x = "sample", y = expression(micrometers^2))

# Plot corrected area of folding per embryo
ggplot(total, aes(Genotype, AreaCorr, colour = Genotype)) + geom_violin(draw_quantiles = c("0.5"), show.legend = FALSE) + geom_jitter(alpha = 0.8, width = 0.1, show.legend = FALSE) + labs(x = "mutant", y = expression(micrometers^2)) + facet_wrap(~Stock) + scale_color_manual(values=c("#0072b2", "#e69f03")) + theme_classic()
ggsave("folded-area_all_facet.png")

# Plot correlation between raw area and number of folds
ggplot(total, aes(Folds, Area, colour = Sample)) + geom_point()
# Plot correlation between corrected area and number of folds
ggplot(total, aes(Folds, AreaCorr, colour = Sample)) + geom_point()

```

