---
title: "Variation of cell area after mitosis"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: February 11, 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Paper/1-analyses/cell-area-post-mitosis")
```

## Goal and Approach

Do epithelial cells occupy a larger area after mitosis? How much area is added
to the surface after cell divisions?

I measured the apical area of individual cells anterior to the cephalic furrow
at three time points: before gastrulation, immediately before cell division
(maximum area), and after mitosis (exactly 10 frames = ~9 min).

The dataset is the cartographic projection of a heterozygote btd-gap embryo
(E10, wild type embryogenesis).

## Load data Experiment metadata

I edited the `Results.txt` file in a spreadsheet to facilitate plotting before
loading the data.

```{r load}
library(knitr)
library(ggplot2)
results <- read.csv("CELLAREA_E10.csv")
```

## Variation in the apical area of individual cells

```{r cellarea}
library(ggbeeswarm)
set.seed(123)
#ggplot(results, aes(Time, Area, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0)
#ggplot(results, aes(Time, Perim, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0)

ggplot(results, aes(Time, Area, colour = Lineage)) +
  geom_quasirandom(size = 4, width = 0.3, alpha = 0.8, show.legend = FALSE) +
  scale_y_continuous(labels = function(x) x / 1000) +
  labs(x = "time", y = expression(paste("apical area ", (px^2~x~10^3)))) +
  scale_x_discrete(labels = c("T1", "T2", "T3")) +
  theme_classic() +
  theme(text =  element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1),
        )
ggsave("apical-area-per-cell.png", width = 10, height = 5)
ggsave("apical-area-per-cell.svg", width = 10, height = 5)
```

## Relative variation in the apical area of individual cells

```{r relativearea}
ggplot(results, aes(Time, NormArea, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0) + ylim(0,3)
#ggplot(results, aes(Time, NormPerim, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0) + ylim(0,3)
```

## Variation in the total apical area per cell lineage

```{r totalarea}
ggplot(results, aes(Time, TotalArea, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0)
#ggplot(results, aes(Time, TotalPerim, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0)
```

## Relative variation in the total apical area per cell lineage

```{r relativetotalarea}
ggplot(results, aes(Time, NormTotalArea, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0)
#ggplot(results, aes(Time, NormTotalPerim, colour = Lineage)) + geom_jitter(alpha = 0.8, width = 0.1, height = 0)
```

## Summary

```{r summary}
rows <- c("Individual Cell Area", "Total Cell Area", "Relative Cell Area", "Relative Total Area")
T1 <- c(mean(results[results$Time == "t1", "Area"]),
        mean(results[results$Time == "t1", "TotalArea"]),
        mean(results[results$Time == "t1", "NormArea"]),
        mean(results[results$Time == "t1", "NormTotalArea"]))
T2 <- c(mean(results[results$Time == "t2", "Area"]),
        mean(results[results$Time == "t2", "TotalArea"]),
        mean(results[results$Time == "t2", "NormArea"]),
        mean(results[results$Time == "t2", "NormTotalArea"]))
T3 <- c(mean(results[results$Time == "t3", "Area"]),
        mean(results[results$Time == "t3", "TotalArea"]),
        mean(results[results$Time == "t3", "NormArea"]),
        mean(results[results$Time == "t3", "NormTotalArea"]))
sumtable <- data.frame(T1, T2, T3, row.names = rows)
kable(round(sumtable[1:2,], digits = 0))
kable(round(sumtable[3:4,], digits = 2))
```
