---
title: "Mitotic expansion relative to ectopic folding"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: September 16, 2024
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/mitotic-expansion")
library(ggplot2)
library(ggbeeswarm)
```

## Import measurements

We exported the measurements of individual embryos using the "Measure" command from Fiji. Each file contains 2 rows with predefined values. Row 1 is the position of the germ band at the frame where the first mitotic expansion happens. Row 2 is the position of the germ band at the frame where the first mitotic cell divides. The X and Y values are in microns and the Slice column corresponds to the frame.

Here is a sample file:

```         
 	Label	X	Y	Slice
1	3D_btd-gap2_z3_t55s_E7.tif:0046-0106-1336:45	398	30	42
2	3D_btd-gap2_z3_t55s_E7.tif:0046-0106-1336:45	365	30	46
```

These raw files need to be parsed, processed, and compiled in a single data frame before plotting. The next block will create an initial data frame.

```{r import-measurements}

# List all txt files in ROIs folder
files = list.files(path = "1-rois", pattern="*.tsv", full.names = TRUE)

# Function for pre-processing
process <- function(tabfile) {
  
  # Read tabular file
  df <- read.delim(tabfile)
  
  # Get label
  label_md <- df$Label[1]
  
  # Clean up embryo name
  embryo <- strsplit(df$Label[1], ":")[[1]][1]
  embryo <- gsub("3D_", "", embryo)
  embryo <- gsub(".tif", "", embryo)
  
  # Get X positions
  x_md_start <- df$X[1]
  x_md_split <- df$X[2]
  
  # Get frames
  frame_md_start <- df$Slice[1]
  frame_md_split <- df$Slice[2]
  
  # Save as vector
  mv <- c(label_md = label_md,
          embryo = embryo,
          x_md_start = x_md_start,
          x_md_split = x_md_split,
          frame_md_start = frame_md_start,
          frame_md_split = frame_md_split)
  
  return(mv)
}

# Define empty data frame
md_expansion <- data.frame()

# Loop over files binding rows to data frame
for (tabfile in files) {
  md_row <- process(tabfile)
  md_expansion <- rbind(md_expansion, as.data.frame(t(md_row)))
}

# Save data frame to file
write.csv(md_expansion, file = "md_expansion.csv", row.names = FALSE)

```

## Merge datasets

Next, we need to import the master dataset file to get some metadata and the germband-extension data for the size and buckling information. The most important values are the stock, zygosity, timestep, and gastrulation frame.

```{r merge-datasets}

# Import main dataset
datasets <- read.csv("../../0-data/datasets.csv")

# Import germband-extension dataset
gb_extension <- read.csv("../germband-extension/gb_extension.csv")
germband <- read.csv("../germband-extension/germband.csv")
gb_extension <- cbind(gb_extension, germband[, c("time_tf_buck", "time_tf_deep")])

# Merge mitotic landmarks with datasets
mitotic <- merge(gb_extension, md_expansion, by = "embryo", all.x = TRUE)
mitotic <- merge(mitotic, datasets, by = "embryo", all.x = TRUE)

# Properly convert column types before any calculations
mitotic <- as.data.frame(lapply(mitotic, type.convert, as.is = FALSE))

```

## Calculate time and percentages

Now that the datasets are merged, we need to convert the frame information to time relative to gastrulation and the position information to percentage of extension.

```{r calculate-time-percentage}

# Calculate GB relative position
mitotic$gb_pos_md_start <- mitotic$x_md_start - mitotic$x_ant
mitotic$gb_pos_md_split <- mitotic$x_md_split - mitotic$x_ant

# Calculate GB percentage extension (%)
mitotic$gb_pct_md_start <- 100 - ((mitotic$gb_pos_md_start / mitotic$size) * 100)
mitotic$gb_pct_md_split <- 100 - ((mitotic$gb_pos_md_split / mitotic$size) * 100)

# Calculate MD time after gastrulation (min)
mitotic$time_md_start <- ((mitotic$frame_md_start - mitotic$frame_gastr) * mitotic$timestep) / 60
mitotic$time_md_split <- ((mitotic$frame_md_split - mitotic$frame_gastr) * mitotic$timestep) / 60

```

## Create long format data frame

For plotting, it is best to reshape the data frame to a long format.

```{r reshape-long-format}

# Define variables for plotting
longvars <- c(
  "embryo",
  "stock",
  "zygosity",
  "size",
  "gb_pct_tf_buck",
  "gb_pct_md_start",
  "gb_pct_md_split",
  "time_tf_buck",
  "time_md_start",
  "time_md_split"
)

mdlong <- reshape(
  mitotic[, longvars],
  varying = list(
    c(
      "gb_pct_tf_buck",
      "gb_pct_md_start",
      "gb_pct_md_split"
    ),
    c(
      "time_tf_buck",
      "time_md_start",
      "time_md_split"
    )
  ),
  v.names = c("gb_pct", "time"),
  timevar = "status",
  times = c("tf_buck", "md_start", "md_split"),
  direction = "long"
)

# Remove the id column that reshape adds
mdlong$id <- NULL

# Transform data frame to order the stock for plotting
mdlong$stock = factor(mdlong$stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
mdlong$zygosity = factor(mdlong$zygosity, levels = c("wildtype", "heterozygote", "homozygote", "double heterozygote", "btd homozygote", "stg homozygote", "double homozygote"))
mdlong$status = factor(mdlong$status, levels = c("tf_buck", "tf_deep", "md_start", "md_split"))
mdlong <- mdlong[order(mdlong$stock, mdlong$zygosity),]
rownames(mdlong) <- 1:nrow(mdlong)

# Save both data frames to file
write.csv(mitotic, file = "mitotic.csv", row.names = FALSE)
write.csv(mdlong, file = "mdlong.csv", row.names = FALSE)

```

## Calculate mean and standard deviation

Create data frame with descriptive statistics for all the different samples and variables.

```{r calculate-mean-sd}
# Automatically aggregate data by columns for MD statuses only
stats <- aggregate(
  cbind(gb_pct, time) ~ stock + zygosity + status,
  mdlong[mdlong$status %in% c("md_start", "md_split"), ],
  function(x) c(
    count = length(x),
    mean = mean(x),
    sd = sd(x))
  )

# Makes column names nicer
stats <- do.call(data.frame, stats)

# Order rows by something meaningful (and re-index)
stats <- stats[order(stats$status, stats$stock, stats$zygosity),]
rownames(stats) <- 1:nrow(stats)

# Print and save data frame
stats
write.csv(stats, file = "stats.csv", row.names = FALSE)

```

## Exploratory plots

For `md_start` (initial mitotic expansion) and `md_split` (first cell division).

### Time of mitotic events after gastrulation by stock (cols) and status (rows)

The plot suggests that mitotic expansion and division is delayed in btd and eve homozygotes and eve heterozygotes, and is probably faster in prd heterozygotes and homozygotes.

```{r plot-md-time-by-stock-status}
ggplot(mdlong[mdlong$status %in% c("md_start", "md_split"), ], aes(stock, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values = c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values = c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "time after gastrulation (min)") +
  facet_grid(rows = vars(status), cols = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size = 1))
ggsave("2-plots/md_time_all_by_stock_status.png", width = 15, height = 7)
ggsave("2-plots/md_time_all_by_stock_status.svg", width = 15, height = 7)
```

### Percentage of germ band extension by stock (cols) and status (rows)

The plot suggests that mitotic expansion and division happens is delayed relative to germ band extension in btd homozygotes.

```{r plot-gb-pct-by-stock-status}
ggplot(mdlong[mdlong$status %in% c("md_start", "md_split"),], aes(stock, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "germ band extension (%)") +
  facet_grid(rows = vars(status), cols = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_all_by_stock_status.png", width = 15, height = 7)
ggsave("2-plots/gb_pct_all_by_stock_status.svg", width = 15, height = 7)
```

### Time of mitotic events after gastrulation by status (cols) and stock (rows)

The plot suggests that buckling in btd, eve, and prd homozygotes happen after the initial mitotic expansion, in contrast to the respective heterozygotes.

```{r plot-md-time-by-status-stock}
ggplot(mdlong, aes(status, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  #geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "time after gastrulation (min)") +
  #facet_grid(cols = vars(zygosity), scales = "free") +
  facet_grid(cols = vars(status), rows = vars(stock), scales = "free_x") +
  #facet_grid(rows = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/md_time_all_by_status_stock.png", width = 10, height = 12)
ggsave("2-plots/md_time_all_by_status_stock.svg", width = 10, height = 12)
```
### Percentage of germ band extension by status (cols) and stock (rows)

The plot suggests that buckling in btd, eve, and prd homozygotes is delayed relative to the initial mitotic expansion, in contrast to the respective heterozygotes.

```{r plot-gb-pct-by-status-stock}
ggplot(mdlong, aes(status, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  #geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "germ band extension (%)") +
  #facet_grid(cols = vars(zygosity), scales = "free") +
  facet_grid(cols = vars(status), rows = vars(stock), scales = "free_x") +
  #facet_grid(rows = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_all_by_status_stock.png", width = 10, height = 12)
ggsave("2-plots/gb_pct_all_by_status_stock.svg", width = 10, height = 12)
```

## Main plots

Comparison between `tf_buck` and `md_start`, the most informative values, excluding data from `prd` since cephalic furrow formation is not prevented in homozygote embryos.

### Time of mitotic expansion vs ectopic folding by status (cols)

The plot shows how ectopic folding is delayed relative to the cephalic furrow, and how the timing of mitotic expansion does not change between different zygosities. Comparing the timing of ectopic folding with mitotic expansion in homozygotes indicates that the latter precedes the former.

```{r plot-md-time-tf-buck-vs-md-start-by-status}
ggplot(mdlong[(mdlong$status %in% c("tf_buck", "md_start") & mdlong$stock != "prd"),], aes(zygosity, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 5)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "zygosity", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(status), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/md_time_tf_buck_vs_md_start_by_status.png", width = 10, height = 6)
ggsave("2-plots/md_time_tf_buck_vs_md_start_by_status.svg", width = 10, height = 6)
```
### Percentage of germ band extension at mitotic expansion vs ectopic folding by status (cols)

Although heterozygote cephalic furrow is slightly delayed compared to wildtype, it happens at the same percentage of germ band extension. This suggests that germ band extension is a good relative measure for developmental events.

This plot shows that ectopic folding is very much delayed relative to germ band extension in homozygotes, and mitotic expansion is unchanged. This means that mitotic expansion precedes ectopic buckling.

```{r plot-gb-pct-tf-buck-vs-md-start-by-status}
ggplot(mdlong[(mdlong$status %in% c("tf_buck", "md_start") & mdlong$stock != "prd"),], aes(zygosity, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "zygosity", y = "germ band extension (%)") +
  facet_grid(cols = vars(status), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_tf_buck_vs_md_start_by_status.png", width = 10, height = 6)
ggsave("2-plots/gb_pct_tf_buck_vs_md_start_by_status.svg", width = 10, height = 6)
```
### Time of mitotic expansion vs ectopic folding by zygosity (cols)

While the previous plots are informative, it is important to determine if the mitotic expansion precedes ectopic folding per individual embryos. To visualize we changed the facet to zygosity and connect the dots of the same embryos.

The plot shows that in heterozygotes, cephalic furrow formation always precedes mitotic expansion, but that in homozygotes it's the reverse, mitotic expansion always precedes ectopic folding.

```{r plot-md-time-tf-buck-vs-md-start-by-zygosity}
ggplot(mdlong[(mdlong$status %in% c("tf_buck", "md_start") & mdlong$stock != "prd"),], aes(status, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE) +
  geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(zygosity), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/md_time_tf_buck_vs_md_start_by_zygosity.png", width = 12, height = 5)
ggsave("2-plots/md_time_tf_buck_vs_md_start_by_zygosity.svg", width = 12, height = 5)
```


### Percentage of germ band extension at mitotic expansion vs ectopic folding by zygosity (cols)


```{r plot-gb-pct-tf-buck-vs-md-start-by-zygosity}
ggplot(mdlong[(mdlong$status %in% c("tf_buck", "md_start") & mdlong$stock != "prd"),], aes(status, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE) +
  geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "germ band extension (%)") +
  facet_grid(cols = vars(zygosity), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_tf_buck_vs_md_start_by_zygosity.png", width = 12, height = 5)
ggsave("2-plots/gb_pct_tf_buck_vs_md_start_by_zygosity.svg", width = 12, height = 5)
```

## Statistical tests

```{r tests}
# Create data frame for tests
tests <- data.frame()

# Contrasts of single variables within stocks (het vs hom)
for (stock in unique(mdlong$stock)) {
  
  if (stock == "wt") { next }
  
  for (status in c("md_start", "md_split")) {
    
    for (value in c("gb_pct", "time")) {
      
      het = mdlong[(mdlong$status == status & mdlong$stock == stock & mdlong$zygosity == "heterozygote"), value]
      hom = mdlong[(mdlong$status == status & mdlong$stock == stock & mdlong$zygosity == "homozygote"), value]
      
      test = wilcox.test(het, hom, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(het) - mean(hom)
      
      test_row = c(
        contrast = "het vs hom",
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.6f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
    }
  }
}

# Contrasts between variables within zygosity (buck vs md)
for (stock in unique(mdlong$stock)) {
  
  for (zygosity in unique(mdlong$zygosity)) {
    
    for (value in c("gb_pct", "time")) {
      
      if (stock == "wt" & zygosity != "wildtype") { next }
      
      if (stock != "wt" & zygosity == "wildtype") { next }
      
      # Note: with low n and all values from one contrast higher or lower than the other, pvalues might be identical
      buck = mdlong[(mdlong$status == "tf_buck" & mdlong$stock == stock & mdlong$zygosity == zygosity), value]
      md = mdlong[(mdlong$status == "md_start" & mdlong$stock == stock & mdlong$zygosity == zygosity), value]
      
      test = wilcox.test(buck, md, paired = TRUE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(buck) - mean(md)
      
      test_row = c(
        contrast = "buck vs md",
        stock = stock,
        zygosity = zygosity,
        status = "-",
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.6f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = TRUE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
    }
  }
}

# Contrasts between variables within zygosity but batch stocks (buck vs md)
for (zygosity in c("heterozygote", "homozygote")) {
  
  for (value in c("gb_pct", "time")) {
    
    # Note: prd excluded because CF formation is not prevented
    buck = mdlong[(mdlong$status == "tf_buck" & mdlong$stock != "prd" & mdlong$zygosity == zygosity), value]
    md = mdlong[(mdlong$status == "md_start" & mdlong$stock != "prd" & mdlong$zygosity == zygosity), value]
    
    test = wilcox.test(buck, md, paired = TRUE)
    pvalue = test$p.value
    statistic = test$statistic
    delta = mean(buck) - mean(md)
    
    test_row = c(
      contrast = "buck vs md",
      stock = "btd+eve",
      zygosity = zygosity,
      status = "-",
      value = value,
      delta = sprintf("%.1f", delta),
      significant = pvalue < 0.05,
      pvalue = sprintf("%.6f", pvalue),
      statistic = sprintf("%s=%s", names(statistic), statistic),
      paired = TRUE
    )
    
    tests <- rbind(tests, as.data.frame(t(test_row)))
  }
}

# Define factors for ordering
tests$contrast = factor(tests$contrast, levels = c("het vs hom", "buck vs md"))
tests$stock = factor(tests$stock, levels = c("wt", "btd", "eve", "prd", "btd+eve"))
tests$status = factor(tests$status, levels = c("md_start", "md_split", "-"))
tests$value = factor(tests$value, levels = c("time", "gb_pct"))
tests <- tests[order(tests$contrast, tests$stock, tests$status, tests$value),]

tests
write.csv(tests, file = "tests.csv", row.names = FALSE)
```
