---
title: "Analysis of evil twins experiment"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: July 23, 2024
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = getwd())
library(ggplot2)
library(ggbeeswarm)
library(svglite)
```

## Experimental setup

Mount RGap embryos on 25 positions with two conditions, control and evil twins. The control embryos are alone, this is to guarantee that they are not affected by their neighbors. The evil twins has the subject with two other embryos with their anterior tip pointed to the local where the cephalic furrow usually forms at the head-trunk boundary. The embryos are mounted in Halocarbon 200 inside a hypoxia chamber. The oxygen levels are put to 10% for about 2h before and after gastrulation.

## Data description

There are two data sources. One CSV file with the technical information of individual datasets, as well as annotations describing the frame in which each developmental event occurs in individual embryos. The other dataset consists of coordinates (xy values in microns) with the position of anatomical landmarks in each embryo.

## Import developmental events

```{r import-events}
# Import datasets with annotated developmental traits
evil_twins <- read.csv("annotations.csv")
```

## Define developmental events

```{r define-events}
events <- c("experiment_start",
            "experiment_end",
            "hypoxia_start",
            "hypoxia_end",
            "cf_folding_start",
            "gb_extension_start",
            "gb_extension_end",
            "gb_extension_total",
            "clypeolabrum_start",
            "gb_retraction_start",
            "gb_retraction_end",
            "gb_retraction_total",
            "head_involution_start",
            "head_involution_end",
            "head_involution_total",
            "dorsal_closure_start",
            "dorsal_closure_end",
            "dorsal_closure_total",
            "gut_partition_start",
            "gut_partition_end",
            "gut_partition_total",
            "twitching_start",
            "tracheae_start",
            "tracheae_to_hatching_total",
            "hatching")
```

## Convert frames to developmental time

```{r process-events}
# Loop over events to calculate time relative to gastrulation
for (event in events) {
  
  # Skip events with total
  if (grepl("total", event)) { next }

  # Define frame and time variable names
  frame_event <- sprintf("frame_%s", event)
  time_event <- sprintf("time_%s", event)
  
  # Create new column with time of events in hours relative to gastrulation
  evil_twins[,time_event] <- round((((evil_twins[,frame_event] - evil_twins$frame_gastrulation) * evil_twins$timestep) + ifelse(evil_twins[,frame_event] > evil_twins$frame_hypoxia_end, evil_twins$time_plucking, 0)) / 60, digits = 2)
  
  # Add experiment total duration time in hours
  if (event == "experiment_end") {
    evil_twins$time_experiment_total <- round(
      ((evil_twins$frame_experiment_end * evil_twins$timestep) + evil_twins$time_plucking) / 60,
      digits = 2)
  }
  
  # Add hypoxia total duration time in hours
  if (event == "hypoxia_end") {
    evil_twins$time_hypoxia_total <- round(
      ((evil_twins$frame_hypoxia_end - evil_twins$frame_hypoxia_start) * evil_twins$timestep) / 60,
      digits = 2)
    
    # Add a time gastrulation before other time variables
    evil_twins$time_gastrulation <- 0
  }
  
  # Add germband extension total duration time in hours
  if (event == "gb_extension_end") {
    evil_twins$time_gb_extension_total <- round(
      (((evil_twins$frame_gb_extension_end - evil_twins$frame_gb_extension_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_gb_extension_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_gb_extension_end"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add germband retraction total duration time in hours
  if (event == "gb_retraction_end") {
    evil_twins$time_gb_retraction_total <- round(
      (((evil_twins$frame_gb_retraction_end - evil_twins$frame_gb_retraction_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_gb_retraction_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_gb_retraction_end"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }

  # Add head involution total duration time in hours
  if (event == "head_involution_end") {
    evil_twins$time_head_involution_total <- round(
      (((evil_twins$frame_head_involution_end - evil_twins$frame_head_involution_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_head_involution_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_head_involution_end"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }

  # Add dorsal closure total duration time in hours
  if (event == "dorsal_closure_end") {
    evil_twins$time_dorsal_closure_total <- round(
      (((evil_twins$frame_dorsal_closure_end - evil_twins$frame_dorsal_closure_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_dorsal_closure_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_dorsal_closure_end"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add dorsal closure total duration time in hours
  if (event == "dorsal_closure_end") {
    evil_twins$time_dorsal_closure_total <- round(
      (((evil_twins$frame_dorsal_closure_end - evil_twins$frame_dorsal_closure_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_dorsal_closure_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_dorsal_closure_end"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add gut partition total duration time in hours
  if (event == "gut_partition_end") {
    evil_twins$time_gut_partition_total <- round(
      (((evil_twins$frame_gut_partition_end - evil_twins$frame_gut_partition_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_gut_partition_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_gut_partition_end"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add gut partition total duration time in hours
  if (event == "tracheae_start") {
    evil_twins$time_tracheae_to_hatching_total <- round(
      (((evil_twins$frame_hatching - evil_twins$frame_tracheae_start) * evil_twins$timestep) + ifelse((evil_twins[,"frame_tracheae_start"] < evil_twins$frame_hypoxia_end & evil_twins[,"frame_hatching"] > evil_twins$frame_hypoxia_end), evil_twins$time_plucking, 0)) / 60,
      digits = 2)
  }

}
```

## Build data frame with full experiments overview

```{r build-overview}
# Build dataframe with aggregated data
overview <- data.frame(
  experiment = as.vector(unique(evil_twins$experiment)),
  date = as.vector(unique(evil_twins$date)),
  total_n = as.vector(table(evil_twins$experiment)),
  total_bad_embryo = as.vector(tapply(evil_twins$bad_embryo, evil_twins$experiment, sum, na.rm = TRUE)),
  total_bad_twins = as.vector(tapply(evil_twins$bad_twins, evil_twins$experiment, sum, na.rm = TRUE)),
  total_hatched = as.vector(tapply(evil_twins$hatched, evil_twins$experiment, sum, na.rm = TRUE)),
  total_time_hypoxia = as.vector(unique(na.omit(round(evil_twins$time_hypoxia_total, digits = 2)))),
  total_time_experiment = as.vector(unique(evil_twins$time_experiment_total))
)

# Add columns with percentage of bad and hatched embryos
overview$pct_bad_embryo <- round((overview$total_bad_embryo / overview$total_n) * 100, digits = 2)
overview$pct_hatched <- round((overview$total_hatched / (overview$total_n - overview$total_bad_embryo)) * 100, digits = 2)

# Calculate mean and sd for each event
for (event in events) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  control_mean <- sprintf("%s_control_mean", time_event)
  control_sd <- sprintf("%s_control_sd", time_event)
  evil_twins_mean <- sprintf("%s_evil_twins_mean", time_event)
  evil_twins_sd <- sprintf("%s_evil_twins_sd", time_event)
  
  # Handy function to apply mean or sd for each condition
  apply_mean_sd <- function(data, condition, func) {
    as.vector(round(tapply(
      data[data$condition == condition, time_event],
      data[data$condition == condition, "experiment"],
      func, na.rm = TRUE
      ), digits = 2))
  }
  
  # Calculate mean and sd values per condition
  overview[,control_mean] <- apply_mean_sd(evil_twins, "control", mean)
  overview[,control_sd] <- apply_mean_sd(evil_twins, "control", sd)
  overview[,evil_twins_mean] <- apply_mean_sd(evil_twins, "evil_twins", mean)
  overview[,evil_twins_sd] <- apply_mean_sd(evil_twins, "evil_twins", sd)
  
}
```

## Calculate normalized values using control mean

```{r calculate-norm-full}
for (event in events[5:length(events)]) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  time_norm <- sprintf("%s_norm", time_event)
  control_mean <- sprintf("%s_control_mean", time_event)
  
  # Create evil_twins-sized vector with control means per experiment
  control_mean_subset <- merge(evil_twins[, c("experiment", "date", "position", time_event)], overview[, c("experiment", control_mean)], by = "experiment", all.x = TRUE)
  
  # Order properly just in case before creating vector
  control_mean_subset[order(control_mean_subset$date, control_mean_subset$position),]
  control_mean_vector <- control_mean_subset[, control_mean]
  
  # Calculate normalized values using control mean
  evil_twins[,time_norm] <- evil_twins[,time_event] - control_mean_vector
  # Divide by control mean to get ratio
  #evil_twins[,time_norm] <- (evil_twins[,time_event] - control_mean_vector) / control_mean_vector
}
```

## Calculate percentage values relative developmental time

```{r calculate-pct}
# Add percentage of developmental time for gastrulation
evil_twins$time_gastrulation_pct <- 0

# Add percentage for the remaining variables
for (event in events[5:length(events)]) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  time_pct <- sprintf("%s_pct", time_event)
  
  # Calculate percentage of developmental time using hatching time
  evil_twins[,time_pct] <- (evil_twins[,time_event] / evil_twins$time_hatching) * 100
}
```

## Hatched embryos vs bad embryos per experiment

```{r pct-hatched}
ggplot(overview, aes(pct_bad_embryo, pct_hatched, color = date)) +
  geom_point(alpha = 0.8, size = 5, show.legend = TRUE) + 
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/overview-pct_bad_embryo-vs-pct_hatched.png", width = 10, height = 7)
#ggsave("2-plots/overview-pct_bad_embryo-vs-pct_hatched.svg", width = 10, height = 7)

```

## Function plot hypoxia starting time vs other events

```{r function-plot-hypoxia}
plot_time_hypoxia_start_vs <- function(data, time_event, prefix = "", faceted = TRUE, plot_width = 25, plot_height = 8, cex = 1) {
  
  # Create scatter plot with points
  plot <- ggplot(data, aes(time_hypoxia_start, data[,time_event], size = time_hypoxia_total, color = condition)) +
    geom_point(alpha = 0.8, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = cex)) + 
    scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
    scale_color_manual(values=c("#56b4e9", "#e69f03")) +
    scale_size(range = c(3,5)) +
    labs(x = "time_hypoxia_start (h)", y = sprintf("%s (h)", time_event)) +
    theme_classic(base_family = "DejaVu Sans") +
    theme(
      text = element_text(size = 35),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 10),
      strip.background = element_blank(),
      panel.background = element_rect(colour = "black", linewidth = 1)
      )
  
  # Add facets if requested
  if (faceted == TRUE) {
    plot <- plot + facet_grid(cols = vars(date), scales = "fixed")
  }
  
  # Define prefix or data name
  if (prefix == "") {
    prefix = deparse(substitute(data))
  }
  
  # Define PNG and SVG output files
  outpng = sprintf("2-plots/%s-time_hypoxia_start-vs-%s.png", prefix, time_event)
  outsvg = sprintf("2-plots/%s-time_hypoxia_start-vs-%s.svg", prefix, time_event)
  ggsave(outpng, width = plot_width, height = plot_height)
  #ggsave(outsvg, width = plot_width, height = plot_height)
  
  # Return needed to output plot
  return(plot)
}
```

## Function plot condition vs other events

```{r function-plot-condition}
plot_condition_vs <- function(data, time_event, prefix = "", faceted = TRUE, plot_width = 25, plot_height = 8, cex = 6) {
  
  # Create scatter plot with points
  plot <- ggplot(data, aes(condition, data[,time_event], fill = condition)) +
    geom_boxplot(aes(color = condition), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
    geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = cex)) +
    scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
    scale_color_manual(values=c("#56b4e9", "#e69f03")) +
    labs(x = "condition", y = sprintf("%s (h)", time_event)) +
    theme_classic(base_family = "DejaVu Sans") +
    theme(text = element_text(size = 35),
          strip.background = element_blank(),
          panel.background = element_rect(colour = "black", linewidth = 1)
          )
  
  # Add facets if requested
  if (faceted == TRUE) {
    plot <- plot + facet_grid(cols = vars(time_hypoxia_total), scales = "free")
  }
  
  # Define prefix or data name
  if (prefix == "") {
    prefix = deparse(substitute(data))
  }
  
  # Define PNG and SVG output files
  outpng = sprintf("2-plots/%s-condition-vs-%s.png", prefix, time_event)
  outsvg = sprintf("2-plots/%s-condition-vs-%s.svg", prefix, time_event)
  ggsave(outpng, width = plot_width, height = plot_height)
  #ggsave(outsvg, width = plot_width, height = plot_height)
  
  # Return needed to output plot
  return(plot)
}
```

## Function plot loop

```{r function-plotloop}
plotloop <- function(items, data, plotfunc, type = "time", faceted = TRUE, plot_width = 25, plot_height = 8, cex = 1) {
  
  # Get dataset name
  dataname = deparse(substitute(data))
  
  # Loop over variables to be plotted
  for (event in items) {
    
    # Coord or time?
    if (type == "coord") {
      type_event <- sprintf("coord_%s", event)
    } else {
      type_event <- sprintf("time_%s", event)
    }
    type_event_norm <- sprintf("%s_norm", type_event)
    type_event_pct <- sprintf("%s_pct", type_event)
    print(type_event)
    
    # Do not plot data that does not exist
    count_nas <- sum(is.na(data[,type_event]))
    count_rows <- length(data[,type_event])
    
    # Call out plot functions using get(dataname) to pass the original name forward
    if (count_nas < count_rows) {
      if (type == "time" | type == "coord") {
        outplot <- plotfunc(data, type_event, prefix = dataname, faceted = faceted, plot_width = plot_width, plot_height = plot_height, cex = cex)
      } else if (type == "norm") {
        outplot <- plotfunc(data, type_event_norm, prefix = dataname, faceted = faceted, plot_width = plot_width, plot_height = plot_height, cex = cex)
      } else if (type == "pct") {
        outplot <- plotfunc(data, type_event_pct, prefix = dataname, faceted = faceted, plot_width = plot_width, plot_height = plot_height, cex = cex)
      }
      print(outplot)
    }
  }
}
```

## Plot hypoxia starting time vs other events

```{r plot-hypoxia}
plotloop(events[5:length(events)], evil_twins, plot_time_hypoxia_start_vs, type = "time")
plotloop(events[5:length(events)], evil_twins, plot_time_hypoxia_start_vs, type = "norm")
plotloop(events[5:length(events)], evil_twins, plot_time_hypoxia_start_vs, type = "pct")
```

## Plot condition vs other events

```{r condition-events}
plotloop(events[5:length(events)], evil_twins, plot_condition_vs, type = "time", cex = 6)
plotloop(events[5:length(events)], evil_twins, plot_condition_vs, type = "norm", cex = 6)
plotloop(events[5:length(events)], evil_twins, plot_condition_vs, type = "pct", cex = 6)
```

## Import developmental landmarks

```{r define-landmarks}
landmarks <- c("body_landmarks",
               "gb_start")
```

```{r import-landmarks}

# Define main data frame with coordinates
coordinates <- data.frame()

# Loop over all landmark groups
for (markgroup in landmarks) {
  
  # Get list of CSV files with coordinates of annotated landmarks
  markdir <- sprintf("1-rois/%s", markgroup)
  coordfiles <- list.files(markdir, pattern = "*.csv", full.names = TRUE)
  
  # Import multiple CSV files in a single data frame
  coordgroup <- do.call(rbind, lapply(coordfiles, read.csv))
  colnames(coordgroup) <- c("row", "label", "x", "y", "frame")
  
  # Create additional columns for downstream use
  coordgroup$file <- sapply(strsplit(coordgroup$label, ":"), "[", 1)
  coordgroup$group <- markgroup
  coordgroup$landmark <- sapply(strsplit(coordgroup$label, ":"), "[", 2)
  coordgroup$date <- sapply(strsplit(coordgroup$label, "_"), "[", 2)
  coordgroup$position <- as.numeric(gsub("Pos", "", sapply(strsplit(coordgroup$label, "_"), "[", 14)))
  
  # Reorder columns and sort rows
  coordgroup <- coordgroup[, c("file", "date", "position", "frame", "group", "landmark", "x", "y")]
  coordgroup <- coordgroup[order(coordgroup$date, coordgroup$position), ]
  
  # Add to main coordinates file
  coordinates <- rbind(coordinates, coordgroup)

}

# Rebuild numeric index
rownames(coordinates) <- 1:nrow(coordinates)
```

## Calculate measurements

```{r distance-functions}
# Calculate the euclidean distance between two points
get_euclidean <- function(ax, ay, bx, by) {
  sqrt((ax - bx)^2 + (ay - by)^2)
}

# Parse data and return a named vector
calculate_distance <- function(a, b) {
  df_a <- coordinates[coordinates$landmark == a, ]
  df_b <- coordinates[coordinates$landmark == b, ]
  dist_ab <- get_euclidean(df_a$x, df_a$y, df_b$x, df_b$y)
  names(dist_ab) <- df_a$file
  return(dist_ab)
}
```

```{r define-distance-pairs}
# Pairs of landmarks to calculate euclidean distance
distpairs <- list(
  gb_start_depth = c("gb_infolding", "vitelline_posterior"),
  gb_start_width = c("gb_edge_left", "gb_edge_right"),
  subject_length = c("subject_anterior", "subject_posterior"),
  subject_width = c("subject_left", "subject_right")
)
```

```{r import-distances}
# Define main distances dataset
distances <- data.frame(dataset = evil_twins$dataset)

# Loop over distance pair names
for (pairname in names(distpairs)) {
  print(pairname)
  # Define points
  a <- distpairs[pairname][[1]][1]
  b <- distpairs[pairname][[1]][2]
  # Calculate distance
  dist_ab <- calculate_distance(a, b)
  # Convert to data frame
  df_ab <- as.data.frame(dist_ab)
  df_ab$dataset <- names(dist_ab)
  colnames(df_ab) <- c(sprintf("coord_%s", pairname), "dataset")
  rownames(df_ab) <- 1:nrow(df_ab)
  # Merge distances with main evil_twins data frame
  distances <- merge(distances, df_ab, by = "dataset", all.x = TRUE)
}

# Hack to remove existing coord columns before merging (avoid duplicates)
#evil_twins <- evil_twins[, -grep("coord", colnames(evil_twins))]

# Merge distances with main evil_twins data frame
evil_twins <- merge(evil_twins, distances, by = "dataset", all.x = TRUE)

# Sort to maintain chronological order
evil_twins <- evil_twins[order(evil_twins$date, evil_twins$position), c(2, 1, 3:ncol(evil_twins))]

# Rebuild numeric index
rownames(evil_twins) <- 1:nrow(evil_twins)
```

## Calculate GB angle

```{r function-angle}
# Get acute angle between lines a-b and c-d
calculate_acute_angle <- function(a, b, c, d) {

  # Reduce subsets to common datasets
  datasets <- Reduce(intersect, list(a$file, b$file, c$file, d$file))
  a <- a[a$file %in% datasets,]
  b <- b[b$file %in% datasets,]
  c <- c[c$file %in% datasets,]
  d <- d[d$file %in% datasets,]

  # Calculate direction vectors
  vector_ab <- cbind(b$x - a$x, b$y - a$y)
  vector_cd <- cbind(d$x - c$x, d$y - c$y)
  
  # Calculate dot product
  dot_product <- vector_ab[,1] * vector_cd[,1] + vector_ab[,2] * vector_cd[,2]
  
  # Calculate magnitudes
  mag_ab <- sqrt(vector_ab[,1]^2 + vector_ab[,2]^2)
  mag_cd <- sqrt(vector_cd[,1]^2 + vector_cd[,2]^2)
  
  # Calculate cosine of angle
  cos_theta <- dot_product / (mag_ab * mag_cd)
  
  # Calculate angle in radians
  theta <- acos(cos_theta)
  
  # Convert to degrees
  angle_deg <- theta * 180 / pi
  
  # Ensure acute angle
  angle_deg <- ifelse(angle_deg > 90, 180 - angle_deg, angle_deg)
  
  # Give names to vector values
  names(angle_deg) <- datasets
  
  return(angle_deg)
}

# Calculate gb_start_angle
angles <- calculate_acute_angle(
    coordinates[coordinates$landmark == "subject_anterior", ],
    coordinates[coordinates$landmark == "subject_posterior", ],
    coordinates[coordinates$landmark == "gb_infolding", ],
    coordinates[coordinates$landmark == "vitelline_posterior", ]
)

# Convert to data frame
df_angles <- as.data.frame(angles)
df_angles$dataset <- names(angles)
colnames(df_angles) <- c("coord_gb_start_angle", "dataset")
rownames(df_angles) <- 1:nrow(df_angles)

# Merge distances with main evil_twins data frame
evil_twins <- merge(evil_twins, df_angles, by = "dataset", all.x = TRUE)

# Sort to maintain chronological order
evil_twins <- evil_twins[order(evil_twins$date, evil_twins$position), c(2, 1, 3:ncol(evil_twins))]

# Rebuild numeric index
rownames(evil_twins) <- 1:nrow(evil_twins)
```


## Condition vs gb starting time per experiment

```{r hatching-experiment}
plotloop(names(distpairs), evil_twins, plot_time_hypoxia_start_vs, type = "coord")
plotloop(names(distpairs), evil_twins, plot_condition_vs, type = "coord", cex = 6)

plotloop(c("gb_start_angle"), evil_twins, plot_time_hypoxia_start_vs, type = "coord")
plotloop(c("gb_start_angle"), evil_twins, plot_condition_vs, type = "coord", cex = 6)
```

## Save full datasets to file

```{r save-full-datasets}
# Save datasets to file
write.csv(evil_twins, file = "evil_twins_full.csv", row.names = FALSE)
write.csv(overview, file = "overview_full.csv", row.names = FALSE)
write.csv(coordinates, file = "coordinates_full.csv", row.names = FALSE)
```

## Discard datasets and embryos below quality thresholds

```{r discard}
# Copy data to clean data frame
dataclean <- evil_twins
# Discard experiments with hatching rate smaller than 90%
dataclean <- dataclean[!dataclean$experiment %in% overview[overview$pct_hatched < 90, ]$experiment, ]
# Discard experiment with abnormal distribution
#dataclean <- dataclean[!dataclean$experiment %in% overview[overview$time_hatching_evil_twins_sd > 1, ]$experiment, ]
# Discard embryos that did not hatch
dataclean <- dataclean[dataclean$hatched, ]
# Discard unfertilized eggs and embryos with catastrophic development
dataclean <- dataclean[!dataclean$bad_embryo, ]
# Discard embryos with bad evil twins (unfertilized or badly positioned)
dataclean <- dataclean[!dataclean$bad_twins, ]
# Discard embryos that were too old
dataclean <- dataclean[!dataclean$time_experiment_start >= 0, ]
```

## Build data frame with cleaned experiments overview

```{r build-overclean}
# Build dataframe with aggregated data
overclean <- data.frame(
  experiment = as.vector(unique(dataclean$experiment)),
  date = as.vector(unique(dataclean$date)),
  total_n = as.vector(table(dataclean$experiment)),
  total_bad_embryo = as.vector(tapply(dataclean$bad_embryo, dataclean$experiment, sum, na.rm = TRUE)),
  total_bad_twins = as.vector(tapply(dataclean$bad_twins, dataclean$experiment, sum, na.rm = TRUE)),
  total_hatched = as.vector(tapply(dataclean$hatched, dataclean$experiment, sum, na.rm = TRUE)),
  total_time_hypoxia = as.vector(unique(na.omit(round(dataclean$time_hypoxia_total, digits = 2)))),
  total_time_experiment = as.vector(unique(dataclean$time_experiment_total))
)

# Add columns with percentage of bad and hatched embryos
overclean$pct_bad_embryo <- round((overclean$total_bad_embryo / overclean$total_n) * 100, digits = 2)
overclean$pct_hatched <- round((overclean$total_hatched / (overclean$total_n - overclean$total_bad_embryo)) * 100, digits = 2)

# Calculate mean and sd for each event
for (event in events) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  control_mean <- sprintf("%s_control_mean", time_event)
  control_sd <- sprintf("%s_control_sd", time_event)
  evil_twins_mean <- sprintf("%s_evil_twins_mean", time_event)
  evil_twins_sd <- sprintf("%s_evil_twins_sd", time_event)
  
  # Handy function to apply mean or sd for each condition
  apply_mean_sd <- function(data, condition, func) {
    as.vector(round(tapply(
      data[data$condition == condition, time_event],
      data[data$condition == condition, "experiment"],
      func, na.rm = TRUE
      ), digits = 2))
  }
  
  # Calculate mean and sd values per condition
  overclean[,control_mean] <- apply_mean_sd(dataclean, "control", mean)
  overclean[,control_sd] <- apply_mean_sd(dataclean, "control", sd)
  overclean[,evil_twins_mean] <- apply_mean_sd(dataclean, "evil_twins", mean)
  overclean[,evil_twins_sd] <- apply_mean_sd(dataclean, "evil_twins", sd)
  
}
```

## Calculate clean normalized values using control mean

```{r calculate-norm-clean}
for (event in events[5:length(events)]) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  time_norm <- sprintf("%s_norm", time_event)
  control_mean <- sprintf("%s_control_mean", time_event)
  
  # Create evil_twins-sized vector with control means per experiment
  control_mean_subset <- merge(dataclean[, c("experiment", "date", "position", time_event)], overclean[, c("experiment", control_mean)], by = "experiment", all.x = TRUE)
  
  # Order properly just in case before creating vector
  control_mean_subset[order(control_mean_subset$date, control_mean_subset$position),]
  control_mean_vector <- control_mean_subset[, control_mean]
  
  # Calculate normalized values using control mean
  dataclean[,time_norm] <- dataclean[,time_event] - control_mean_vector
}
```

## Save cleaned datasets to file

```{r save-cleaned-datasets}
# Save datasets to file
write.csv(dataclean, file = "evil_twins_clean.csv", row.names = FALSE)
write.csv(overclean, file = "overview_clean.csv", row.names = FALSE)
```

## Plot cleaned hypoxia starting time vs other events

```{r time-start-hypoxia}
plotloop(events[5:length(events)], dataclean, plot_time_hypoxia_start_vs, type = "time", faceted = FALSE, plot_width = 13, plot_height = 10, cex = 1)
```

## Plot cleaned condition vs other events

```{r time-events-condition}
plotloop(events[5:length(events)], dataclean, plot_condition_vs, type = "norm", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)
plotloop(events[5:length(events)], dataclean, plot_condition_vs, type = "pct", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)
```

## Plot cleaned condition vs other coord measurements

```{r coord-condition-hypoxia-events}
plotloop(names(distpairs), dataclean, plot_time_hypoxia_start_vs, type = "coord", faceted = FALSE, plot_width = 13, plot_height = 10, cex = 1)
plotloop(names(distpairs), dataclean, plot_condition_vs, type = "coord", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)

plotloop(c("gb_start_angle"), dataclean, plot_time_hypoxia_start_vs, type = "coord", faceted = FALSE, plot_width = 13, plot_height = 10, cex = 1)
plotloop(c("gb_start_angle"), dataclean, plot_condition_vs, type = "coord", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)
```


## Create long format data frame

For plotting the comparison of developmental events.

```{r reshape-long-format}

# Define variables for plotting
longvars <- c(
  "date",
  "dataset",
  "position",
  "condition",
  "time_gastrulation",
  "time_cf_folding_start",
  "time_gb_extension_start",
  "time_gb_extension_end",
  "time_gb_retraction_start",
  "time_gb_retraction_end",
  "time_head_involution_end",
  "time_gut_partition_start",
  "time_tracheae_start",
  "time_hatching",
  "time_gastrulation_pct",
  "time_cf_folding_start_pct",
  "time_gb_extension_start_pct",
  "time_gb_extension_end_pct",
  "time_gb_retraction_start_pct",
  "time_gb_retraction_end_pct",
  "time_head_involution_end_pct",
  "time_gut_partition_start_pct",
  "time_tracheae_start_pct",
  "time_hatching_pct"
)

datalong <- reshape(
  dataclean[, longvars],
  varying = list(
    c(
      "time_gastrulation",
      "time_cf_folding_start",
      "time_gb_extension_start",
      "time_gb_extension_end",
      "time_gb_retraction_start",
      "time_gb_retraction_end",
      "time_gut_partition_start",
      "time_head_involution_end",
      "time_tracheae_start",
      "time_hatching"
    ),
    c(
      "time_gastrulation_pct",
      "time_cf_folding_start_pct",
      "time_gb_extension_start_pct",
      "time_gb_extension_end_pct",
      "time_gb_retraction_start_pct",
      "time_gb_retraction_end_pct",
      "time_gut_partition_start_pct",
      "time_head_involution_end_pct",
      "time_tracheae_start_pct",
      "time_hatching_pct"
    )
  ),
  v.names = c("time", "pct"),
  timevar = "event",
  times = c(
    "gastrulation",
    "cf_folding_start",
    "gb_extension_start",
    "gb_extension_end",
    "gb_retraction_start",
    "gb_retraction_end",
    "gut_partition_start",
    "head_involution_end",
    "tracheae_start",
    "hatching"
  ),
  direction = "long"
)

# Remove the id column that reshape adds
datalong$id <- NULL

# Remove rows with NA
datalong <- na.omit(datalong)

# Transform data frame to order the stock for plotting
datalong$condition <- factor(datalong$condition, levels = c("control", "evil_twins"))
datalong$event <- factor(
  datalong$event,
  levels = c(
    "gastrulation",
    "cf_folding_start",
    "gb_extension_start",
    "gb_extension_end",
    "gb_retraction_start",
    "gb_retraction_end",
    "gut_partition_start",
    "head_involution_end",
    "tracheae_start",
    "hatching"
  )
)
datalong <- datalong[order(datalong$date, datalong$event, datalong$condition),]
rownames(datalong) <- 1:nrow(datalong)

# Save long data frame to file
write.csv(datalong, file = "evil_twins_clean_long.csv", row.names = FALSE)

```

## Plot timing of developmental events

```{r plot-timing-events}
ggplot(datalong, aes(time, event, color = event, fill = event, group = interaction(event, condition))) +
  geom_boxplot(show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 1.0) +
  labs(x = "time (h)", y = "event") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/datalong-time_event-vs-condition.png", width = 10, height = 4)
ggsave("2-plots/datalong-time_event-vs-condition.svg", width = 10, height = 4)
```


## Plot percentage of developmental events

```{r plot-percentage-events}
ggplot(datalong, aes(pct, event, color = event, fill = event, group = interaction(event, condition))) +
  geom_boxplot(show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 1.0) +
  labs(x = "time (h)", y = "event") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/datalong-pct_event-vs-condition.png", width = 10, height = 4)
ggsave("2-plots/datalong-pct_event-vs-condition.svg", width = 10, height = 4)
```

## Plot percentage of developmental events (bar plot)

```{r plot-percentage-events-bar}
ggplot(datalong[datalong$date %in% c("2024-07-25", "2024-07-29") & datalong$time >= 0, ], aes(time, dataset, fill = event)) +
  geom_bar(position = position_fill(reverse = TRUE), stat = "identity") +
  labs(x = "time (h)", y = "event") +
  facet_grid(rows = vars(condition), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        axis.text.y=element_blank(),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/datalong-bar-pct_event-vs-condition.png", width = 10, height = 4)
ggsave("2-plots/datalong-bar-pct_event-vs-condition.svg", width = 10, height = 4)
```

## Calculate mean and standard deviation

Create data frame from clean dataset with descriptive statistics for different conditions and variables.

```{r values-clean}

# Generate list with columns of interest
nonempty_colnames <- colnames(dataclean[, colSums(!is.na(dataclean)) > 0])
stats_colnames <- nonempty_colnames[36:length(nonempty_colnames)]

# Create empty dataframe for descriptive stats
stats <- data.frame(
  condition = c("control", "evil_twins")
)

# Automatically aggregate data by columns since each column has a different number of rows with values
for (column in stats_colnames) {
  
  # Calculate stats for column (column ~ condition)
  col_stats <- aggregate(
    reformulate("condition", column),
    dataclean,
    function(x)
    c(
      count = length(x),
      mean = mean(x),
      sd = sd(x)
    ))
  
  # Bind columns with main data frame
  stats <- merge(stats, col_stats, by="condition")
  
}

# Makes column names nicer
stats <- do.call(data.frame, stats)

# Print and save data frame
stats
write.csv(stats, file = "stats.csv", row.names = FALSE)

```

## Statistics

```{r stats}
tests <- data.frame()

for (column in stats_colnames) {
  
  control <- dataclean[dataclean$condition == "control", column]
  evil <- dataclean[dataclean$condition == "evil_twins", column]
  
  
  # Control vs Evil Twins
  contrast <- "control vs evil twins"
  test <- wilcox.test(control, evil, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(control) - mean(evil)
  
  test_row <- c(
    variable = column,
    contrast = contrast,
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests <- rbind(tests, as.data.frame(t(test_row)))

}

tests
write.csv(tests, file = "tests.csv", row.names = FALSE)
```

## Select representative datasets

Select datasets that are closest to the normalized mean values for representing the data in figures.

```{r representative-datasets}
# Get mean for time_hatching_norm
# Check control embryos closest to the mean (~0.0)
# Check evil twins closest to the mean (~1.0)

time_hatching_norm_mean_control <- stats[stats$condition == "control", "time_hatching_norm.mean"]
time_hatching_norm_mean_evil <- stats[stats$condition == "evil_twins", "time_hatching_norm.mean"]

controls_time_hatching_norm <- dataclean[dataclean$condition == "control", c("dataset", "date", "time_hatching_norm")]
controls_time_hatching_norm$distance_to_norm <- abs(time_hatching_norm_mean_control - abs(controls_time_hatching_norm$time_hatching_norm))
controls_time_hatching_norm <- controls_time_hatching_norm[order(controls_time_hatching_norm$distance_to_norm), ]
rownames(controls_time_hatching_norm) <- 1:nrow(controls_time_hatching_norm)
controls_time_hatching_norm


evils_time_hatching_norm <- dataclean[dataclean$condition == "evil_twins", c("dataset", "date", "time_hatching_norm")]
evils_time_hatching_norm$distance_to_norm <- abs(time_hatching_norm_mean_evil - abs(evils_time_hatching_norm$time_hatching_norm))
evils_time_hatching_norm <- evils_time_hatching_norm[order(evils_time_hatching_norm$distance_to_norm), ]
rownames(evils_time_hatching_norm) <- 1:nrow(evils_time_hatching_norm)
evils_time_hatching_norm
```

