---
title: "Analysis of evil twins experiment"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: July 23, 2024
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/evil-twins")
library(ggplot2)
library(ggbeeswarm)
library(svglite)
```

## Experimental setup

Mount RGap embryos on 25 positions with two conditions, control and evil twins. The control embryos are alone, this is to guarantee that they are not affected by their neighbors. The evil twins has the subject with two other embryos with their anterior tip pointed to the local where the cephalic furrow usually forms at the head-trunk boundary. The embryos are mounted in Halocarbon 200 inside a hypoxia chamber. The oxygen levels are put to 10% for about 2h before and after gastrulation.

## Data description

There are two data sources. One CSV file with the technical information of individual datasets, as well as annotations describing the frame in which each developmental event occurs in individual embryos. The other dataset consists of coordinates (xy values in microns) with the position of anatomical landmarks in each embryo.

## Import developmental events

```{r import-events}
# Import datasets with annotated developmental traits
datafull <- read.csv("evil_twins.csv")
```

## Define developmental events

```{r define-events}
events <- c("experiment_start",
            "experiment_end",
            "hypoxia_start",
            "hypoxia_end",
            "cf_folding_start",
            "gb_extension_start",
            "gb_extension_end",
            "gb_extension_total",
            "clypeolabrum_start",
            "gb_retraction_start",
            "gb_retraction_end",
            "gb_retraction_total",
            "head_involution_start",
            "head_involution_end",
            "head_involution_total",
            "dorsal_closure_start",
            "dorsal_closure_end",
            "dorsal_closure_total",
            "gut_partition_start",
            "gut_partition_end",
            "gut_partition_total",
            "twitching_start",
            "tracheae_start",
            "tracheae_to_hatching_total",
            "hatching")
```

## Convert frames to developmental time

```{r process-events}
# Loop over events to calculate time relative to gastrulation
for (event in events) {
  
  # Skip events with total
  if (grepl("total", event)) { next }

  # Define frame and time variable names
  frame_event <- sprintf("frame_%s", event)
  time_event <- sprintf("time_%s", event)
  
  # Create new column with time of events in hours relative to gastrulation
  datafull[,time_event] <- round((((datafull[,frame_event] - datafull$frame_gastrulation) * datafull$timestep) + ifelse(datafull[,frame_event] > datafull$frame_hypoxia_end, datafull$time_plucking, 0)) / 60, digits = 2)
  
  # Add experiment total duration time in hours
  if (event == "experiment_end") {
    datafull$time_experiment_total <- round(
      ((datafull$frame_experiment_end * datafull$timestep) + datafull$time_plucking) / 60,
      digits = 2)
  }
  
  # Add hypoxia total duration time in hours
  if (event == "hypoxia_end") {
    datafull$time_hypoxia_total <- round(
      ((datafull$frame_hypoxia_end - datafull$frame_hypoxia_start) * datafull$timestep) / 60,
      digits = 2)
  }
  
  # Add germband extension total duration time in hours
  if (event == "gb_extension_end") {
    datafull$time_gb_extension_total <- round(
      (((datafull$frame_gb_extension_end - datafull$frame_gb_extension_start) * datafull$timestep) + ifelse((datafull[,"frame_gb_extension_start"] < datafull$frame_hypoxia_end & datafull[,"frame_gb_extension_end"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add germband retraction total duration time in hours
  if (event == "gb_retraction_end") {
    datafull$time_gb_retraction_total <- round(
      (((datafull$frame_gb_retraction_end - datafull$frame_gb_retraction_start) * datafull$timestep) + ifelse((datafull[,"frame_gb_retraction_start"] < datafull$frame_hypoxia_end & datafull[,"frame_gb_retraction_end"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }

  # Add head involution total duration time in hours
  if (event == "head_involution_end") {
    datafull$time_head_involution_total <- round(
      (((datafull$frame_head_involution_end - datafull$frame_head_involution_start) * datafull$timestep) + ifelse((datafull[,"frame_head_involution_start"] < datafull$frame_hypoxia_end & datafull[,"frame_head_involution_end"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }

  # Add dorsal closure total duration time in hours
  if (event == "dorsal_closure_end") {
    datafull$time_dorsal_closure_total <- round(
      (((datafull$frame_dorsal_closure_end - datafull$frame_dorsal_closure_start) * datafull$timestep) + ifelse((datafull[,"frame_dorsal_closure_start"] < datafull$frame_hypoxia_end & datafull[,"frame_dorsal_closure_end"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add dorsal closure total duration time in hours
  if (event == "dorsal_closure_end") {
    datafull$time_dorsal_closure_total <- round(
      (((datafull$frame_dorsal_closure_end - datafull$frame_dorsal_closure_start) * datafull$timestep) + ifelse((datafull[,"frame_dorsal_closure_start"] < datafull$frame_hypoxia_end & datafull[,"frame_dorsal_closure_end"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add gut partition total duration time in hours
  if (event == "gut_partition_end") {
    datafull$time_gut_partition_total <- round(
      (((datafull$frame_gut_partition_end - datafull$frame_gut_partition_start) * datafull$timestep) + ifelse((datafull[,"frame_gut_partition_start"] < datafull$frame_hypoxia_end & datafull[,"frame_gut_partition_end"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }
  
  # Add gut partition total duration time in hours
  if (event == "tracheae_start") {
    datafull$time_tracheae_to_hatching_total <- round(
      (((datafull$frame_hatching - datafull$frame_tracheae_start) * datafull$timestep) + ifelse((datafull[,"frame_tracheae_start"] < datafull$frame_hypoxia_end & datafull[,"frame_hatching"] > datafull$frame_hypoxia_end), datafull$time_plucking, 0)) / 60,
      digits = 2)
  }

}
```

## Build data frame with full experiments overview

```{r build-overview}
# Build dataframe with aggregated data
overfull <- data.frame(
  experiment = as.vector(unique(datafull$experiment)),
  date = as.vector(unique(datafull$date)),
  total_n = as.vector(table(datafull$experiment)),
  total_bad_embryo = as.vector(tapply(datafull$bad_embryo, datafull$experiment, sum, na.rm = TRUE)),
  total_bad_twins = as.vector(tapply(datafull$bad_twins, datafull$experiment, sum, na.rm = TRUE)),
  total_hatched = as.vector(tapply(datafull$hatched, datafull$experiment, sum, na.rm = TRUE)),
  total_time_hypoxia = as.vector(unique(na.omit(round(datafull$time_hypoxia_total, digits = 2)))),
  total_time_experiment = as.vector(unique(datafull$time_experiment_total))
)

# Add columns with percentage of bad and hatched embryos
overfull$pct_bad_embryo <- round((overfull$total_bad_embryo / overfull$total_n) * 100, digits = 2)
overfull$pct_hatched <- round((overfull$total_hatched / (overfull$total_n - overfull$total_bad_embryo)) * 100, digits = 2)

# Calculate mean and sd for each event
for (event in events) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  control_mean <- sprintf("%s_control_mean", time_event)
  control_sd <- sprintf("%s_control_sd", time_event)
  evil_twins_mean <- sprintf("%s_evil_twins_mean", time_event)
  evil_twins_sd <- sprintf("%s_evil_twins_sd", time_event)
  
  # Handy function to apply mean or sd for each condition
  apply_mean_sd <- function(data, condition, func) {
    as.vector(round(tapply(
      data[data$condition == condition, time_event],
      data[data$condition == condition, "experiment"],
      func, na.rm = TRUE
      ), digits = 2))
  }
  
  # Calculate mean and sd values per condition
  overfull[,control_mean] <- apply_mean_sd(datafull, "control", mean)
  overfull[,control_sd] <- apply_mean_sd(datafull, "control", sd)
  overfull[,evil_twins_mean] <- apply_mean_sd(datafull, "evil_twins", mean)
  overfull[,evil_twins_sd] <- apply_mean_sd(datafull, "evil_twins", sd)
  
}
```

## Calculate normalized values using control mean

```{r calculate-norm}
for (event in events[5:length(events)]) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  time_norm <- sprintf("%s_norm", time_event)
  control_mean <- sprintf("%s_control_mean", time_event)
  
  # Create datafull-sized vector with control means per experiment
  control_mean_subset <- merge(datafull[, c("experiment", "date", "position", time_event)], overfull[, c("experiment", control_mean)], by = "experiment", all.x = TRUE)
  
  # Order properly just in case before creating vector
  control_mean_subset[order(control_mean_subset$date, control_mean_subset$position),]
  control_mean_vector <- control_mean_subset[, control_mean]
  
  # Calculate normalized values using control mean
  datafull[,time_norm] <- datafull[,time_event] - control_mean_vector
  # Divide by control mean to get ratio
  #datafull[,time_norm] <- (datafull[,time_event] - control_mean_vector) / control_mean_vector
}
```

## Calculate percentage values relative developmental time

```{r calculate-pct}
for (event in events[5:length(events)]) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  time_pct <- sprintf("%s_pct", time_event)
  
  # Calculate normalized values using control mean
  datafull[,time_pct] <- (datafull[,time_event] / datafull$time_hatching) * 100
}
```

## Hatched embryos vs bad embryos per experiment

```{r pct-hatched}
ggplot(overfull, aes(pct_bad_embryo, pct_hatched, color = date)) +
  geom_point(alpha = 0.8, size = 5, show.legend = TRUE) + 
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/overfull-pct_bad_embryo-vs-pct_hatched.png", width = 10, height = 7)
#ggsave("2-plots/overfull-pct_bad_embryo-vs-pct_hatched.svg", width = 10, height = 7)

```

## Function plot hypoxia starting time vs other events

```{r function-plot-hypoxia}
plot_time_hypoxia_start_vs <- function(data, time_event, prefix = "", faceted = TRUE, plot_width = 25, plot_height = 8, cex = 1) {
  
  # Create scatter plot with points
  plot <- ggplot(data, aes(time_hypoxia_start, data[,time_event], size = time_hypoxia_total, color = condition)) +
    geom_point(alpha = 0.8, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = cex)) + 
    scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
    scale_color_manual(values=c("#56b4e9", "#e69f03")) +
    scale_size(range = c(3,5)) +
    labs(x = "time_hypoxia_start (h)", y = sprintf("%s (h)", time_event)) +
    theme_classic(base_family = "DejaVu Sans") +
    theme(
      text = element_text(size = 35),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 10),
      strip.background = element_blank(),
      panel.background = element_rect(colour = "black", linewidth = 1)
      )
  
  # Add facets if requested
  if (faceted == TRUE) {
    plot <- plot + facet_grid(cols = vars(date), scales = "fixed")
  }
  
  # Define prefix or data name
  if (prefix == "") {
    prefix = deparse(substitute(data))
  }
  
  # Define PNG and SVG output files
  outpng = sprintf("2-plots/%s-time_hypoxia_start-vs-%s.png", prefix, time_event)
  outsvg = sprintf("2-plots/%s-time_hypoxia_start-vs-%s.svg", prefix, time_event)
  ggsave(outpng, width = plot_width, height = plot_height)
  #ggsave(outsvg, width = plot_width, height = plot_height)
  
  # Return needed to output plot
  return(plot)
}
```

## Function plot condition vs other events

```{r function-plot-condition}
plot_condition_vs <- function(data, time_event, prefix = "", faceted = TRUE, plot_width = 25, plot_height = 8, cex = 6) {
  
  # Create scatter plot with points
  plot <- ggplot(data, aes(condition, data[,time_event], fill = condition)) +
    geom_boxplot(aes(color = condition), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
    geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = cex)) +
    scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
    scale_color_manual(values=c("#56b4e9", "#e69f03")) +
    labs(x = "condition", y = sprintf("%s (h)", time_event)) +
    theme_classic(base_family = "DejaVu Sans") +
    theme(text = element_text(size = 35),
          strip.background = element_blank(),
          panel.background = element_rect(colour = "black", linewidth = 1)
          )
  
  # Add facets if requested
  if (faceted == TRUE) {
    plot <- plot + facet_grid(cols = vars(time_hypoxia_total), scales = "free")
  }
  
  # Define prefix or data name
  if (prefix == "") {
    prefix = deparse(substitute(data))
  }
  
  # Define PNG and SVG output files
  outpng = sprintf("2-plots/%s-condition-vs-%s.png", prefix, time_event)
  outsvg = sprintf("2-plots/%s-condition-vs-%s.svg", prefix, time_event)
  ggsave(outpng, width = plot_width, height = plot_height)
  #ggsave(outsvg, width = plot_width, height = plot_height)
  
  # Return needed to output plot
  return(plot)
}
```

## Function plot loop

```{r function-plotloop}
plotloop <- function(items, data, plotfunc, type = "time", faceted = TRUE, plot_width = 25, plot_height = 8, cex = 1) {
  
  # Get dataset name
  dataname = deparse(substitute(data))
  
  # Loop over variables to be plotted
  for (event in items) {
    
    # Coord or time?
    if (type == "coord") {
      type_event <- sprintf("coord_%s", event)
    } else {
      type_event <- sprintf("time_%s", event)
    }
    type_event_norm <- sprintf("%s_norm", type_event)
    type_event_pct <- sprintf("%s_pct", type_event)
    print(type_event)
    
    # Do not plot data that does not exist
    count_nas <- sum(is.na(data[,type_event]))
    count_rows <- length(data[,type_event])
    
    # Call out plot functions using get(dataname) to pass the original name forward
    if (count_nas < count_rows) {
      if (type == "time" | type == "coord") {
        outplot <- plotfunc(data, type_event, prefix = dataname, faceted = faceted, plot_width = plot_width, plot_height = plot_height, cex = cex)
      } else if (type == "norm") {
        outplot <- plotfunc(data, type_event_norm, prefix = dataname, faceted = faceted, plot_width = plot_width, plot_height = plot_height, cex = cex)
      } else if (type == "pct") {
        outplot <- plotfunc(data, type_event_pct, prefix = dataname, faceted = faceted, plot_width = plot_width, plot_height = plot_height, cex = cex)
      }
      print(outplot)
    }
  }
}
```

## Plot hypoxia starting time vs other events

```{r plot-hypoxia}
plotloop(events[5:length(events)], datafull, plot_time_hypoxia_start_vs, type = "time")
plotloop(events[5:length(events)], datafull, plot_time_hypoxia_start_vs, type = "norm")
plotloop(events[5:length(events)], datafull, plot_time_hypoxia_start_vs, type = "pct")
```


## Plot condition vs other events

```{r condition-events}
plotloop(events[5:length(events)], datafull, plot_condition_vs, type = "time", cex = 6)
plotloop(events[5:length(events)], datafull, plot_condition_vs, type = "norm", cex = 6)
plotloop(events[5:length(events)], datafull, plot_condition_vs, type = "pct", cex = 6)
```

## Import developmental landmarks

```{r define-landmarks}
landmarks <- c("body_landmarks",
               "gb_start")
```

```{r import-landmarks}

# Define main data frame with coordinates
coordinates <- data.frame()

# Loop over all landmark groups
for (markgroup in landmarks) {
  
  # Get list of CSV files with coordinates of annotated landmarks
  markdir <- sprintf("1-rois/%s", markgroup)
  coordfiles <- list.files(markdir, pattern = "*.csv", full.names = TRUE)
  
  # Import multiple CSV files in a single data frame
  coordgroup <- do.call(rbind, lapply(coordfiles, read.csv))
  colnames(coordgroup) <- c("row", "label", "x", "y", "frame")
  
  # Create additional columns for downstream use
  coordgroup$file <- sapply(strsplit(coordgroup$label, ":"), "[", 1)
  coordgroup$group <- markgroup
  coordgroup$landmark <- sapply(strsplit(coordgroup$label, ":"), "[", 2)
  coordgroup$date <- sapply(strsplit(coordgroup$label, "_"), "[", 2)
  coordgroup$position <- as.numeric(gsub("Pos", "", sapply(strsplit(coordgroup$label, "_"), "[", 14)))
  
  # Reorder columns and sort rows
  coordgroup <- coordgroup[, c("file", "date", "position", "frame", "group", "landmark", "x", "y")]
  coordgroup <- coordgroup[order(coordgroup$date, coordgroup$position), ]
  
  # Add to main coordinates file
  coordinates <- rbind(coordinates, coordgroup)

}

# Rebuild numeric index
rownames(coordinates) <- 1:nrow(coordinates)
```

## Calculate measurements

```{r distance-functions}
# Calculate the euclidean distance between two points
get_euclidean <- function(ax, ay, bx, by) {
  sqrt((ax - bx)^2 + (ay - by)^2)
}

# Parse data and return a named vector
calculate_distance <- function(a, b) {
  df_a <- coordinates[coordinates$landmark == a, ]
  df_b <- coordinates[coordinates$landmark == b, ]
  dist_ab <- get_euclidean(df_a$x, df_a$y, df_b$x, df_b$y)
  names(dist_ab) <- df_a$file
  return(dist_ab)
}
```

```{r define-distance-pairs}
# Pairs of landmarks to calculate euclidean distance
distpairs <- list(
  gb_start_depth = c("gb_infolding", "vitelline_posterior"),
  gb_start_width = c("gb_edge_left", "gb_edge_right"),
  subject_length = c("subject_anterior", "subject_posterior"),
  subject_width = c("subject_left", "subject_right")
)
```

```{r import-distances}
# Define main distances dataset
distances <- data.frame(dataset = datafull$dataset)

# Loop over distance pair names
for (pairname in names(distpairs)) {
  print(pairname)
  # Define points
  a <- distpairs[pairname][[1]][1]
  b <- distpairs[pairname][[1]][2]
  # Calculate distance
  dist_ab <- calculate_distance(a, b)
  # Convert to data frame
  df_ab <- as.data.frame(dist_ab)
  df_ab$dataset <- names(dist_ab)
  colnames(df_ab) <- c(sprintf("coord_%s", pairname), "dataset")
  rownames(df_ab) <- 1:nrow(df_ab)
  # Merge distances with main datafull data frame
  distances <- merge(distances, df_ab, by = "dataset", all.x = TRUE)
}

# Hack to remove existing coord columns before merging (avoid duplicates)
#datafull <- datafull[, -grep("coord", colnames(datafull))]

# Merge distances with main datafull data frame
datafull <- merge(datafull, distances, by = "dataset", all.x = TRUE)

# Sort to maintain chronological order
datafull <- datafull[order(datafull$date, datafull$position), c(2, 1, 3:ncol(datafull))]

# Rebuild numeric index
rownames(datafull) <- 1:nrow(datafull)
```

## Calculate GB angle

```{r function-angle}
# Get acute angle between lines a-b and c-d
calculate_acute_angle <- function(a, b, c, d) {

  # Reduce subsets to common datasets
  datasets <- Reduce(intersect, list(a$file, b$file, c$file, d$file))
  a <- a[a$file %in% datasets,]
  b <- b[b$file %in% datasets,]
  c <- c[c$file %in% datasets,]
  d <- d[d$file %in% datasets,]

  # Calculate direction vectors
  vector_ab <- cbind(b$x - a$x, b$y - a$y)
  vector_cd <- cbind(d$x - c$x, d$y - c$y)
  
  # Calculate dot product
  dot_product <- vector_ab[,1] * vector_cd[,1] + vector_ab[,2] * vector_cd[,2]
  
  # Calculate magnitudes
  mag_ab <- sqrt(vector_ab[,1]^2 + vector_ab[,2]^2)
  mag_cd <- sqrt(vector_cd[,1]^2 + vector_cd[,2]^2)
  
  # Calculate cosine of angle
  cos_theta <- dot_product / (mag_ab * mag_cd)
  
  # Calculate angle in radians
  theta <- acos(cos_theta)
  
  # Convert to degrees
  angle_deg <- theta * 180 / pi
  
  # Ensure acute angle
  angle_deg <- ifelse(angle_deg > 90, 180 - angle_deg, angle_deg)
  
  # Give names to vector values
  names(angle_deg) <- datasets
  
  return(angle_deg)
}

# Calculate gb_start_angle
angles <- calculate_acute_angle(
    coordinates[coordinates$landmark == "subject_anterior", ],
    coordinates[coordinates$landmark == "subject_posterior", ],
    coordinates[coordinates$landmark == "gb_infolding", ],
    coordinates[coordinates$landmark == "vitelline_posterior", ]
)

# Convert to data frame
df_angles <- as.data.frame(angles)
df_angles$dataset <- names(angles)
colnames(df_angles) <- c("coord_gb_start_angle", "dataset")
rownames(df_angles) <- 1:nrow(df_angles)

# Merge distances with main datafull data frame
datafull <- merge(datafull, df_angles, by = "dataset", all.x = TRUE)

# Sort to maintain chronological order
datafull <- datafull[order(datafull$date, datafull$position), c(2, 1, 3:ncol(datafull))]

# Rebuild numeric index
rownames(datafull) <- 1:nrow(datafull)
```


## Condition vs gb starting time per experiment

```{r hatching-experiment}
plotloop(names(distpairs), datafull, plot_time_hypoxia_start_vs, type = "coord")
plotloop(names(distpairs), datafull, plot_condition_vs, type = "coord", cex = 6)

plotloop(c("gb_start_angle"), datafull, plot_time_hypoxia_start_vs, type = "coord")
plotloop(c("gb_start_angle"), datafull, plot_condition_vs, type = "coord", cex = 6)
```

## Discard datasets and embryos below quality thresholds

```{r discard}
# Copy data to clean data frame
dataclean <- datafull
# Discard experiments with hatching rate smaller than 90%
dataclean <- dataclean[!dataclean$experiment %in% overfull[overfull$pct_hatched < 90, ]$experiment, ]
# Discard experiment with abnormal distribution
#dataclean <- dataclean[!dataclean$experiment %in% overfull[overfull$time_hatching_evil_twins_sd > 1, ]$experiment, ]
# Discard embryos that did not hatch
dataclean <- dataclean[dataclean$hatched, ]
# Discard unfertilized eggs and embryos with catastrophic development
dataclean <- dataclean[!dataclean$bad_embryo, ]
# Discard embryos with bad evil twins (unfertilized or badly positioned)
dataclean <- dataclean[!dataclean$bad_twins, ]
# Discard embryos that were too old
dataclean <- dataclean[!dataclean$time_experiment_start >= 0, ]
```

## Build data frame with cleaned experiments overview

```{r build-overview}
# Build dataframe with aggregated data
overclean <- data.frame(
  experiment = as.vector(unique(dataclean$experiment)),
  date = as.vector(unique(dataclean$date)),
  total_n = as.vector(table(dataclean$experiment)),
  total_bad_embryo = as.vector(tapply(dataclean$bad_embryo, dataclean$experiment, sum, na.rm = TRUE)),
  total_bad_twins = as.vector(tapply(dataclean$bad_twins, dataclean$experiment, sum, na.rm = TRUE)),
  total_hatched = as.vector(tapply(dataclean$hatched, dataclean$experiment, sum, na.rm = TRUE)),
  total_time_hypoxia = as.vector(unique(na.omit(round(dataclean$time_hypoxia_total, digits = 2)))),
  total_time_experiment = as.vector(unique(dataclean$time_experiment_total))
)

# Add columns with percentage of bad and hatched embryos
overclean$pct_bad_embryo <- round((overclean$total_bad_embryo / overclean$total_n) * 100, digits = 2)
overclean$pct_hatched <- round((overclean$total_hatched / (overclean$total_n - overclean$total_bad_embryo)) * 100, digits = 2)

# Calculate mean and sd for each event
for (event in events) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  control_mean <- sprintf("%s_control_mean", time_event)
  control_sd <- sprintf("%s_control_sd", time_event)
  evil_twins_mean <- sprintf("%s_evil_twins_mean", time_event)
  evil_twins_sd <- sprintf("%s_evil_twins_sd", time_event)
  
  # Handy function to apply mean or sd for each condition
  apply_mean_sd <- function(data, condition, func) {
    as.vector(round(tapply(
      data[data$condition == condition, time_event],
      data[data$condition == condition, "experiment"],
      func, na.rm = TRUE
      ), digits = 2))
  }
  
  # Calculate mean and sd values per condition
  overclean[,control_mean] <- apply_mean_sd(dataclean, "control", mean)
  overclean[,control_sd] <- apply_mean_sd(dataclean, "control", sd)
  overclean[,evil_twins_mean] <- apply_mean_sd(dataclean, "evil_twins", mean)
  overclean[,evil_twins_sd] <- apply_mean_sd(dataclean, "evil_twins", sd)
  
}
```

## Calculate clean normalized values using control mean

```{r calculate-norm}
for (event in events[5:length(events)]) {
  
  # Define new column names
  time_event <- sprintf("time_%s", event)
  time_norm <- sprintf("%s_norm", time_event)
  control_mean <- sprintf("%s_control_mean", time_event)
  
  # Create datafull-sized vector with control means per experiment
  control_mean_subset <- merge(dataclean[, c("experiment", "date", "position", time_event)], overclean[, c("experiment", control_mean)], by = "experiment", all.x = TRUE)
  
  # Order properly just in case before creating vector
  control_mean_subset[order(control_mean_subset$date, control_mean_subset$position),]
  control_mean_vector <- control_mean_subset[, control_mean]
  
  # Calculate normalized values using control mean
  dataclean[,time_norm] <- dataclean[,time_event] - control_mean_vector
}
```

## Plot cleaned hypoxia starting time vs other events

```{r time-start-hatched}
plotloop(events[5:length(events)], dataclean, plot_time_hypoxia_start_vs, type = "time", faceted = FALSE, plot_width = 13, plot_height = 10, cex = 1)
```

## Plot cleaned condition vs other events

```{r time-start-hatched}
plotloop(events[5:length(events)], dataclean, plot_condition_vs, type = "norm", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)
plotloop(events[5:length(events)], dataclean, plot_condition_vs, type = "pct", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)
```

## Plot cleaned condition vs other coord measurements

```{r time-start-hatched}
plotloop(names(distpairs), dataclean, plot_time_hypoxia_start_vs, type = "coord", faceted = FALSE, plot_width = 13, plot_height = 10, cex = 1)
plotloop(names(distpairs), dataclean, plot_condition_vs, type = "coord", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)

plotloop(c("gb_start_angle"), dataclean, plot_time_hypoxia_start_vs, type = "coord", faceted = FALSE, plot_width = 13, plot_height = 10, cex = 1)
plotloop(c("gb_start_angle"), dataclean, plot_condition_vs, type = "coord", faceted = FALSE, plot_width = 10, plot_height = 10, cex = 4)
```

## Percentage plotting

```{r percentage}

# Create data frame
percent <- rbind(
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "00-time_gastrulation",
             value = 0),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "01-time_cf_folding_start",
             value = dataclean$time_cf_folding_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "02-time_gb_extension_start",
             value = dataclean$time_gb_extension_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "03-time_gb_extension_end",
             value = dataclean$time_gb_extension_end),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "04-time_gb_retraction_start",
             value = dataclean$time_gb_retraction_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "05-time_gb_retraction_end",
             value = dataclean$time_gb_retraction_end),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "06-time_head_involution_start",
             value = dataclean$time_head_involution_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "07-time_head_involution_end",
             value = dataclean$time_head_involution_end),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "08-time_gut_partition_start",
             value = dataclean$time_gut_partition_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "09-time_twitching_start",
             value = dataclean$time_twitching_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "10-time_tracheae_start",
             value = dataclean$time_tracheae_start),
  data.frame(embryo = sprintf("%s_%s", dataclean$date, dataclean$position),
             date = dataclean$date,
             condition = dataclean$condition,
             event = "11-time_hatching",
             value = dataclean$time_hatching)
  )

# Set negative values to zero
percent$value <- ifelse(percent$value < 0, 0, percent$value)

#ggplot(percent, aes(embryo, value, fill = event)) +
ggplot(percent[percent$date %in% c("2024-07-25", "2024-07-29"), ], aes(x=reorder(embryo, value), y=value, fill = event)) +
  geom_bar(position = position_fill(reverse = TRUE), stat = "identity") +
  coord_flip() +
  facet_grid(rows = vars(condition), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/dataclean-percent_event-vs-condition.png", width = 15, height = 8)
#ggsave("2-plots/clean-time_hypoxia_start-vs-time_folding.svg", width = 25, height = 8)
```

## Statistics

```{r stats}
statsloop <- function(type) {
  # type = norm or coord
  variables <- colnames(dataclean[, grepl(type, colnames(dataclean))])
  
  for (variable in variables) {
    
    control <- dataclean[dataclean$condition == "control", variable]
    evil_twins <- dataclean[dataclean$condition == "evil_twins", variable]
    
    count_nas <- sum(is.na(control))
    count_rows <- length(control)
    
    if (count_nas < count_rows) {
      
      test <- wilcox.test(control, evil_twins, exact = FALSE)
      print(sprintf("%s >>> p-value = %s", variable, test["p.value"]))
      
    }
  }
}

sink('tests.txt')
statsloop("_norm")
statsloop("pct")
statsloop("coord")
sink()
```








## Archive

Starting time vs hatching time per experiment

```{r time-start-hatched}
ggplot(datafull, aes(time_hypoxia_start, time_hatching, size = time_hypoxia_total, color = condition)) +
  geom_point(alpha = 0.8, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = 1)) + 
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  scale_size(range = c(3,5)) +
  facet_grid(cols = vars(date), scales = "fixed") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/full-time_start-vs-time_hatching.png", width = 25, height = 8)
#ggsave("2-plots/full-time_start-vs-time_hatching.svg", width = 25, height = 8)
```

## Hatching and cephalic furrow

This is a descriptive plot to understand the correlation between hatching success and cephalic furrow inhibition success against the experiment start time and hypoxia start time.

From a general experience of performing the evil twins experiments, if set too early, the embryos will die, if set too late, the CF will not be inhibited. If that's the case, we should see a defined region in the plot.

```{r cf-timing}
ggplot(dataclean[dataclean$condition == "evil_twins",], aes(time_hypoxia_start, time_gb_retraction_end, color = cephalic_furrow)) +
  geom_point(alpha = 0.8, size = 5, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = 1)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#cccccc") +
  geom_hline(yintercept = c(0, 1.5), linetype = "solid", color = "#cccccc") +
  #annotate("rect", xmin = -1.7, xmax = 0.45, ymin = -1.3, ymax = -0.4, alpha = 0.1, fill = "#cccccc") +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  #labs(x = "time at hypoxia start", y = "hatching (h)") +
  facet_grid(cols = vars(time_hypoxia_total), scales = "fixed") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/dataclean-time_start-vs-time_hypoxia_start.png", width = 15, height = 8)
#ggsave("2-plots/dataclean-time_start-vs-time_hypoxia_start.svg", width = 15, height = 8)
```

## Starting time vs normalized hatching time

We expect that embryos that are on earlier stages when mounted, will take longer to hatch. Therefore, a negative correlation between `time_start` and `time_hatching` when time is hours relative to gastrulation.

The average hatching time is different between experiments. It does not correlate with the developmental time that these embryos start.

```{r time-start-hatched}
ggplot(dataclean, aes(time_experiment_start, time_hatching_norm, size = time_hypoxia_total, color = condition)) +
  geom_point(alpha = 0.8, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = 1)) + 
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  scale_size(range = c(3,5)) +
  #facet_grid(cols = vars(condition), scales = "fixed") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/clean-time_start-vs-time_hatching_norm.png", width = 25, height = 8)
#ggsave("2-plots/clean-time_start-vs-time_hatching_norm.svg", width = 25, height = 8)
```

## Starting hypoxia time vs folding time

We expect that embryos that are on earlier stages when mounted, will take longer to hatch. Therefore, a negative correlation between `time_start` and `time_hatching` when time is hours relative to gastrulation.

The average hatching time is different between experiments. It does not correlate with the developmental time that these embryos start.

```{r time-start-hatched}
ggplot(dataclean, aes(time_hypoxia_start, time_folding_norm, size = time_hypoxia_total, color = condition)) +
  geom_point(alpha = 0.8, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = 1)) + 
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  scale_size(range = c(3,5)) +
  #facet_grid(cols = vars(condition), scales = "fixed") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/clean-time_hypoxia_start-vs-time_folding.png", width = 25, height = 8)
#ggsave("2-plots/clean-time_hypoxia_start-vs-time_folding.svg", width = 25, height = 8)
```


## Starting time vs normalized gb retraction time

```{r time-start-hatched}
ggplot(dataclean, aes(time_experiment_start, time_gb_retraction_end_norm, size = time_hypoxia_total, color = condition)) +
  geom_point(alpha = 0.8, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 0.1, cex = 1)) + 
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  scale_size(range = c(3,5)) +
  #facet_grid(cols = vars(condition), scales = "fixed") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/clean-time_start-vs-time_gb_retraction_norm.png", width = 25, height = 8)
#ggsave("2-plots/clean-time_start-vs-time_gb_retraction_norm.svg", width = 25, height = 8)
```

## Condition vs normalized hatching time

Difference in the time of hatching between control and evil twins.

```{r hatching-experiment}
ggplot(dataclean, aes(condition, time_hatching_norm, fill = condition)) +
  geom_boxplot(aes(color = condition), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  #labs(x = "condition", y = "hatching (h)") +
  #facet_grid(cols = vars(time_hypoxia_total), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1))
ggsave("2-plots/clean-condition-vs-time_hatching_norm.png", width = 25, height = 8)
#ggsave("2-plots/clean-condition-vs-time_hatching_norm.svg", width = 25, height = 8)
```

## Condition vs folding time

```{r hatching-experiment}
ggplot(dataclean, aes(condition, time_folding, fill = condition)) +
  geom_boxplot(aes(color = condition), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  #labs(x = "condition", y = "hatching (h)") +
  #facet_grid(cols = vars(time_hypoxia_total), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1))
ggsave("2-plots/clean-condition-vs-time_folding.png", width = 25, height = 8)
#ggsave("2-plots/clean-condition-vs-time_folding.svg", width = 25, height = 8)
```

## Condition vs normalized gb retraction time

```{r hatching-experiment}
ggplot(dataclean, aes(condition, time_gb_retraction_end_norm, fill = condition)) +
  geom_boxplot(aes(color = condition), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  #labs(x = "condition", y = "hatching (h)") +
  #facet_grid(cols = vars(time_hypoxia_total), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1))
ggsave("2-plots/clean-condition-vs-time_folding.png", width = 25, height = 8)
#ggsave("2-plots/clean-condition-vs-time_folding.svg", width = 25, height = 8)
```


```{r stats}
# Normal if p-value > 0.05
print("Testing normality of distribution")
shapiro.test(dataclean$time_hatching_norm)
shapiro.test(dataclean[dataclean$condition == "control", ]$time_hatching_norm)
shapiro.test(dataclean[dataclean$condition == "evil_twins", ]$time_hatching_norm)

print("Testing for difference between samples")
wilcox.test(dataclean[dataclean$condition == "control", ]$time_hatching_norm,
            dataclean[dataclean$condition == "evil_twins", ]$time_hatching_norm)
wilcox.test(dataclean[dataclean$condition == "control", ]$time_head_involution_end_norm,
            dataclean[dataclean$condition == "evil_twins", ]$time_head_involution_end_norm)

print("t-test just to get the difference between sample means")
t.test(dataclean[dataclean$condition == "control", ]$time_hatching_norm,
       dataclean[dataclean$condition == "evil_twins", ]$time_hatching_norm,
       var.equal = FALSE)

#sink()
```
