---
title: "Measurements of slp mutants"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: February 06, 2020
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/slp-mutants")
library(ggplot2)
library(ggbeeswarm)
```

## Import length measurements

To analyze the differences between control and slp mutants, we manually annotated the egg, trunk, and germ band lengths at the frame in which the initiator cells first appear in lateral 3D renderings.

```{r import-lengths}
slp_lengths <- read.csv("slp_lengths.csv")
```

## Calculate mean and standard deviation

Create data frame with descriptive statistics for all the different samples and variables.

```{r values-lengths}

# Automatically aggregate data by columns
stats_lengths <- aggregate(cbind(position_cf, position_gb) ~ zygosity, slp_lengths, function(x)
  c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
  ))

# Makes column names nicer
stats_lengths <- do.call(data.frame, stats_lengths)

# Order rows by something meaningful (and re-index)
stats_lengths$zygosity <- factor(stats_lengths$zygosity, levels = c("heterozygote", "homozygote"))
stats_lengths <- stats_lengths[order(stats_lengths$zygosity),]
rownames(stats_lengths) <- 1:nrow(stats_lengths)

# Print and save data frame
stats_lengths
write.csv(stats_lengths, file = "stats_lengths.csv", row.names = FALSE)

```

## Plot CF position

```{r plot-cf-position}
ggplot(slp_lengths, aes(zygosity, position_cf, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 5)) +
  scale_y_continuous(limits = c(50,80)) +
  scale_x_discrete(labels = c("+/−", "−/−")) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  labs(x = expression(italic("slp")), y = "egg length (%)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 30),
        title = element_text(size = 25),
        legend.text = element_text(size = 25),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/cf_pct_position.png", width = 4, height = 4)
ggsave("2-plots/cf_pct_position.svg", width = 4, height = 4)
```

## Plot GB position

```{r plot-gb-position}
#TODO: Plot time after gastrulation instead of GB position

ggplot(slp_lengths, aes(zygosity, position_gb, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 5)) +
  scale_y_continuous(limits = c(0,30)) +
  scale_x_discrete(labels = c("+/−", "−/−")) +
  scale_fill_manual(values=c("#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#56b4e9", "#e69f03")) +
  labs(x = expression(italic("slp")), y = "egg length (%)") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 30),
        title = element_text(size = 25),
        legend.text = element_text(size = 25),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/gb_pct_position.png", width = 4, height = 4)
ggsave("2-plots/gb_pct_position.svg", width = 4, height = 4)
```

## Statistics

```{r stats-lengths}
tests_lengths <- data.frame()

for (position in c("position_cf", "position_gb")) {
  
  het <- slp_lengths[slp_lengths$zygosity == "heterozygote", position]
  hom <- slp_lengths[slp_lengths$zygosity == "homozygote", position]
  
  # Heterozygotes vs homozygotes
  contrast <- "het vs hom"
  test <- wilcox.test(het, hom, paired = FALSE)
  pvalue <- test$p.value
  statistic <- test$statistic
  delta <- mean(het) - mean(hom)
  
  test_row <- c(
    contrast = contrast,
    value = position,
    delta = sprintf("%.1f", delta),
    significant = pvalue < 0.05,
    pvalue = sprintf("%.9f", pvalue),
    statistic = sprintf("%s=%s", names(statistic), statistic),
    paired = FALSE
  )
  
  tests_lengths <- rbind(tests_lengths, as.data.frame(t(test_row)))

}

tests_lengths
write.csv(tests_lengths, file = "tests_lengths.csv", row.names = FALSE)
```

## Import row measurements

To quantify the differences in the number of rows between control and slp mutants, we manually counted the cells between eve stripe 1 and 2 in five different rows for each in situ image.

```{r import-rows}
# Load row counts
slp_rows <- read.csv("slp_rows.csv")

# Transform data frame to order the stock for plotting
slp_rows$zygosity <- factor(slp_rows$zygosity, levels = c("wildtype", "heterozygote", "homozygote"))
slp_rows <- slp_rows[order(slp_rows$zygosity),]
rownames(slp_rows) <- 1:nrow(slp_rows)
```

## Calculate mean and standard deviation

Create data frame with descriptive statistics for all the different samples and variables.

```{r values-rows}

# Automatically aggregate data by columns
stats_rows <- aggregate(count ~ zygosity, slp_rows, function(x)
  c(
    count = length(x),
    mean = mean(x),
    sd = sd(x)
  ))

# Makes column names nicer
stats_rows <- do.call(data.frame, stats_rows)

# Order rows by something meaningful (and re-index)
stats_rows$zygosity <- factor(stats_rows$zygosity, levels = c("wildtype", "heterozygote", "homozygote"))
stats_rows <- stats_rows[order(stats_rows$zygosity),]
rownames(stats_rows) <- 1:nrow(stats_rows)

# Print and save data frame
stats_rows
write.csv(stats_rows, file = "stats_rows.csv", row.names = FALSE)

```
## Plot row counts

```{r plot-count-rows}

# Plot
ggplot(slp_rows, aes(zygosity, count, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 2.5)) +
  scale_y_continuous(limits = c(2,10)) +
  scale_x_discrete(labels = c("+/+", "+/−", "−/−")) +
  scale_fill_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7",  "#56b4e9", "#e69f03")) +
  labs(x = expression(italic("slp")), y = "row count") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 30),
        title = element_text(size = 25),
        legend.text = element_text(size = 25),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", linewidth = 1)
        )
ggsave("2-plots/rows_between_eve1_eve2.png", width = 4, height = 4)
ggsave("2-plots/rows_between_eve1_eve2.svg", width = 4, height = 4)
```

## Statistics

```{r stats-rows}
tests_rows <- data.frame()

wt <- slp_rows[slp_rows$zygosity == "wildtype", "count"]
het <- slp_rows[slp_rows$zygosity == "heterozygote", "count"]
hom <- slp_rows[slp_rows$zygosity == "homozygote", "count"]

# Wildtype vs Heterozygotes
contrast <- "wt vs het"
test <- wilcox.test(wt, het, paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
delta <- mean(wt) - mean(het)

test_row <- c(
  contrast = contrast,
  value = "count",
  delta = sprintf("%.1f", delta),
  significant = pvalue < 0.05,
  pvalue = sprintf("%.9f", pvalue),
  statistic = sprintf("%s=%s", names(statistic), statistic),
  paired = FALSE
)

tests_rows <- rbind(tests_rows, as.data.frame(t(test_row)))

# Wildtype vs Homozygotes
contrast <- "wt vs hom"
test <- wilcox.test(wt, hom, paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
delta <- mean(wt) - mean(hom)

test_row <- c(
  contrast = contrast,
  value = "count",
  delta = sprintf("%.1f", delta),
  significant = pvalue < 0.05,
  pvalue = sprintf("%.9f", pvalue),
  statistic = sprintf("%s=%s", names(statistic), statistic),
  paired = FALSE
)

tests_rows <- rbind(tests_rows, as.data.frame(t(test_row)))

# Heterozygotes vs Homozygotes
contrast <- "het vs hom"
test <- wilcox.test(het, hom, paired = FALSE)
pvalue <- test$p.value
statistic <- test$statistic
delta <- mean(het) - mean(hom)

test_row <- c(
  contrast = contrast,
  value = "count",
  delta = sprintf("%.1f", delta),
  significant = pvalue < 0.05,
  pvalue = sprintf("%.9f", pvalue),
  statistic = sprintf("%s=%s", names(statistic), statistic),
  paired = FALSE
)

tests_rows <- rbind(tests_rows, as.data.frame(t(test_row)))

tests_rows
write.csv(tests_rows, file = "tests_rows.csv", row.names = FALSE)
```