---
title: "Germ band extension relative to ectopic folding"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: October 24, 2022
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/germband-extension")
library(ggplot2)
library(ggbeeswarm)
```

## Import measurements

We exported the measurements of individual embryos using the "Measure" command from Fiji. Each file contains 4 rows with predefined values. Row 1 and 2 are the anterior and posterior limits of the vitelline envelope, respectively. The Slice value of these two rows represents the frame at the onset of gastrulation (T=0). Row 3 marks the position of the tip of the germ band when the cephalic furrow or an ectopic fold first buckle in. The Slice correspond to the frame when this happens. Row 4 marks the position of the tip of the germ band when the cephalic furrow or ectopic folds reach maximum depth, immediately before they begin to unfold. The Slice also corresponds to the frame when this happens.

Here is a sample file:

```         
 	Label	X	Y	Slice
1	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	21	102	26
2	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	520	90	26
3	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	354	34	47
4	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	300	33	56
```

For extracting information and plotting, these raw files need to be parsed, processed, and compiled in a single data frame. The next block reads, process, and creates an initial data frame.

```{r import-measurements}
# List all txt files in ROIs folder
files = list.files(path = "1-rois", pattern="*.tsv", full.names = TRUE)

# Function for pre-processing
process <- function(tabfile) {
  
  # Read txt file
  df <- read.delim(tabfile)
  
  # Get label
  label_tf <- df$Label[1]
  
  # Clean up embryo name
  embryo <- strsplit(df$Label[1], ":")[[1]][1]
  embryo <- gsub("3D_", "", embryo)
  embryo <- gsub(".tif", "", embryo)
  
  # Get X positions
  x_ant <- df$X[1]
  x_post <- df$X[2]
  x_tf_buck <- df$X[3]
  x_tf_deep <- df$X[4]
  
  # Get sizes
  size <- x_post - x_ant
  gb_pos_tf_buck <- x_tf_buck - x_ant
  gb_pos_tf_deep <- x_tf_deep - x_ant
  
  # Get percentages
  gb_pct_tf_buck <- 100 - ((gb_pos_tf_buck / size) * 100)
  gb_pct_tf_deep <- 100 - ((gb_pos_tf_deep / size) * 100)
  
  # Get frames
  frame_gastr <- df$Slice[1]
  frame_tf_buck <- df$Slice[3]
  frame_tf_deep <- df$Slice[4]
  
  # Save as vector
  mv <- c(label_tf = label_tf,
          embryo = embryo,
          size = size,
          x_ant = x_ant,
          x_post = x_post,
          x_tf_buck = x_tf_buck,
          x_tf_deep = x_tf_deep,
          gb_pos_tf_buck = gb_pos_tf_buck,
          gb_pos_tf_deep = gb_pos_tf_deep,
          gb_pct_tf_buck = gb_pct_tf_buck,
          gb_pct_tf_deep = gb_pct_tf_deep,
          frame_gastr = frame_gastr,
          frame_tf_buck = frame_tf_buck,
          frame_tf_deep = frame_tf_deep
          )
  
  return(mv)
}

# Define empty data frame
gb_extension <- data.frame()

# Loop over files binding rows to data frame
for (tabfile in files) {
  gb_row <- process(tabfile)
  gb_extension <- rbind(gb_extension, as.data.frame(t(gb_row)))
}

# Save data frame to file
write.csv(gb_extension, file = "gb_extension.csv", row.names = FALSE)
```

## Merge datasets

Next, we need to import the master dataset file to get some metadata. Most important is the stock, zygosity, timestep, and gastrulation frame.

```{r merge-datasets}

# Import main dataset
datasets <- read.csv("../../0-data/datasets.csv")

# Merge landmarks with datasets
germband <- merge(gb_extension, datasets, by = "embryo", all.x = TRUE)

# Properly convert column types before any calculations
germband <- as.data.frame(lapply(germband, type.convert, as.is = FALSE))

```

## Calculate time

Now that the datasets are merged, we need to convert the frame information to time relative to gastrulation and the position information to percentage of extension.

```{r calculate-time}
# Create columns with times (min)
germband$time_tf_buck <- ((germband$frame_tf_buck - germband$frame_gastr) * germband$timestep) / 60
germband$time_tf_deep <- ((germband$frame_tf_deep - germband$frame_gastr) * germband$timestep) / 60
```

## Create long format data frame

Convert the main data frame into the long format for plotting

```{r reshape-long-format}

# Define variables for plotting
longvars <- c(
  "embryo",
  "stock",
  "zygosity",
  "size",
  "gb_pct_tf_buck",
  "gb_pct_tf_deep",
  "time_tf_buck",
  "time_tf_deep"
)

gblong <- reshape(
  germband[, longvars],
  varying = list(
    c(
      "gb_pct_tf_buck",
      "gb_pct_tf_deep"
    ),
    c(
      "time_tf_buck",
      "time_tf_deep"
    )
  ),
  v.names = c("gb_pct", "time"),
  timevar = "status",
  times = c("tf_buck", "tf_deep"),
  direction = "long"
)

# Remove the id column that reshape adds
gblong$id <- NULL

# Transform data frame to order the stock for plotting
gblong$stock = factor(gblong$stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
gblong$zygosity = factor(gblong$zygosity, levels = c("wildtype", "heterozygote", "homozygote", "double heterozygote", "btd homozygote", "stg homozygote", "double homozygote"))
gblong$status = factor(gblong$status, levels = c("tf_buck", "tf_deep"))
gblong <- gblong[order(gblong$stock, gblong$zygosity),]
rownames(gblong) <- 1:nrow(gblong)

# Save both data frames to file
write.csv(germband, file = "germband.csv", row.names = FALSE)
write.csv(gblong, file = "gblong.csv", row.names = FALSE)
```

## Calculate mean and standard deviation

Create data frame with descriptive statistics for all the different samples and variables.


```{r values}

# Automatically aggregate data by columns
stats <- aggregate(
  cbind(gb_pct, time) ~ stock + zygosity + status,
  gblong,
  function(x)
    c(
      count = length(x),
      mean = mean(x),
      sd = sd(x))
  )

# Makes column names nicer
stats <- do.call(data.frame, stats)

# Order rows by something meaningful (and re-index)
stats <- stats[order(stats$status, stats$stock, stats$zygosity),]
rownames(stats) <- 1:nrow(stats)

# Print and save data frame
stats
write.csv(stats, file = "stats.csv", row.names = FALSE)

```

## Exploratory plots

We want to identify the difference between the cephalic furrow and the ectopic folds in the percentage of germ band extension, and the timing of formation after gastrulation.

### Time of mitotic events after gastrulation by stock (cols) and status (rows)

```{r plot-md-time-by-stock-status}
ggplot(gblong, aes(stock, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "time after gastrulation (min)") +
  facet_grid(rows = vars(status), cols = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_time_all_by_stock_status.png", width = 15, height = 7)
ggsave("2-plots/gb_time_all_by_stock_status.svg", width = 15, height = 7)
```


### Percentage of germ band extension by stock (cols) and status (rows)

Plot the germ band extension (%) for homozygotes and heterozygotes at two different moments: initial infolding/buckling (buck) and deepest point (deep).

```{r plot-gb-pct-by-stock-status}
ggplot(gblong, aes(stock, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "germ band extension (%)") +
  facet_grid(rows = vars(status), cols = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_pct_all_by_stock_status.png", width = 15, height = 7)
ggsave("2-plots/gb_pct_all_by_stock_status.svg", width = 15, height = 7)
```

### Time of mitotic events after gastrulation by status (cols) and stock (rows)

```{r plot-gb-time-by-status-stock}
ggplot(gblong, aes(status, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(status), rows = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_all_by_status_stock.png", width = 7, height = 12)
ggsave("2-plots/gb_time_all_by_status_stock.svg", width = 7, height = 12)
```

### Percentage of germ band extension by status (cols) and stock (rows)

```{r plot-gb-pct-by-status-stock}
ggplot(gblong, aes(status, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "germ band extension (%)") +
  facet_grid(cols = vars(status), rows = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_all_by_status_stock.png", width = 7, height = 12)
ggsave("2-plots/gb_pct_all_by_status_stock.svg", width = 7, height = 12)
```

## Main plots

The above are nice descriptive plots. But for the figure, the most interesting is the initial buckling/infolding relative to GB position and time after gastrulation.

### Time after gastrulation at buckling/infolding by stock (cols)

```{r plot-gb-time-tf-buck-stock}
ggplot(gblong[gblong$status == "tf_buck",], aes(stock, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_time_tf_buck_by_stock.png", width = 12, height = 4)
ggsave("2-plots/gb_time_tf_buck_by_stock.svg", width = 12, height = 4)
```

### Percentage of germ band extension at buckling/infolding by stock (cols)

```{r plot-gb-pct-tf-buck-stock}
ggplot(gblong[gblong$status == "tf_buck",], aes(stock, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "germ band extension (%)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1))
ggsave("2-plots/gb_pct_tf_buck_by_stock.png", width = 12, height = 4)
ggsave("2-plots/gb_pct_tf_buck_by_stock.svg", width = 12, height = 4)
```


### Time of ectopic folding by status (cols)

```{r plot-gb-time-tf-buck-by-status}
ggplot(gblong[(gblong$status %in% c("tf_buck") & gblong$stock != "prd"),], aes(zygosity, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "zygosity", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(status), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_tf_buck_by_status.png", width = 6, height = 6)
ggsave("2-plots/gb_time_tf_buck_by_status.svg", width = 6, height = 6)
```

### Percentage of germ band extension at mitotic expansion vs ectopic folding by status (cols)

```{r plot-gb-pct-tf-buck-by-status}
ggplot(gblong[(gblong$status %in% c("tf_buck") & gblong$stock != "prd"),], aes(zygosity, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "zygosity", y = "germ band extension (%)") +
  facet_grid(cols = vars(status), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_tf_buck_by_status.png", width = 6, height = 6)
ggsave("2-plots/gb_pct_tf_buck_by_status.svg", width = 6, height = 6)
```


## Statistical tests

```{r tests}

#TODO: Finish up tests

# Create data frame for tests
tests <- data.frame()

# Contrasts of single variables within stocks (het vs hom)
for (stock in unique(gblong$stock)) {
  
  if (stock == "wt") { next }
  
  for (status in unique(gblong$status)) {
    
    for (value in c("gb_pct", "time")) {
      
      # Get values to be compared
      wt = gblong[(gblong$status == status & gblong$stock == "wt" & gblong$zygosity == "wildtype"), value]
      het = gblong[(gblong$status == status & gblong$stock == stock & gblong$zygosity == "heterozygote"), value]
      hom = gblong[(gblong$status == status & gblong$stock == stock & gblong$zygosity == "homozygote"), value]
      
      # Heterozygotes vs homozygotes
      contrast = "het vs hom"
      test = wilcox.test(het, hom, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(het) - mean(hom)
      
      test_row = c(
        contrast = contrast,
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.9f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
      # Wildtype vs heterozygotes
      contrast = "wt vs het"
      test = wilcox.test(wt, het, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(wt) - mean(het)
      
      test_row = c(
        contrast = contrast,
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.9f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
      # Wildtype vs homozygotes
      contrast = "wt vs hom"
      test = wilcox.test(wt, hom, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(wt) - mean(hom)
      
      test_row = c(
        contrast = contrast,
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.9f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
    }
  }
}



# Contrasts between variables within zygosity but batch stocks (buck vs md)
for (zygosity in c("heterozygote", "homozygote")) {
  
  for (status in unique(gblong$status)) {
    
    for (value in c("gb_pct", "time")) {
      
      # Get values to be compared
      wt = gblong[(gblong$status == status & gblong$stock == "wt" & gblong$zygosity == "wildtype"), value]
      het = gblong[(gblong$status == status & gblong$stock != "prd" & gblong$zygosity == "heterozygote"), value]
      hom = gblong[(gblong$status == status & gblong$stock != "prd" & gblong$zygosity == "homozygote"), value]
      
      # Define custom stock
      stock = "btd+eve"
      
      # Heterozygotes vs homozygotes
      contrast = "het vs hom"
      test = wilcox.test(het, hom, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(het) - mean(hom)
      
      test_row = c(
        contrast = contrast,
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.9f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
      # Wildtype vs heterozygotes
      contrast = "wt vs het"
      test = wilcox.test(wt, het, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(wt) - mean(het)
      
      test_row = c(
        contrast = contrast,
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.9f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
      # Wildtype vs homozygotes
      contrast = "wt vs hom"
      test = wilcox.test(wt, hom, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(wt) - mean(hom)
      
      test_row = c(
        contrast = contrast,
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.9f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
    }
  }
}

tests
write.csv(tests, file = "tests.csv", row.names = FALSE)
```
