---
title: "Germ band extension relative to CF and EF"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: October 24, 2022
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Paper/1-analyses/germband-extension")
library(ggplot2)
```

## Analysis

At which percentage of egg length (and how long after gastrulation) do ectopic
folds form in cephalic furrow mutants? We have a rough estimate based on a few
embryos but this can be quantified more accurately. A proper characterization of
this dynamics in mutants is important for the model.

## Directories

- `0-data`: 3D renderings of lateral views
- `1-rois`: Landmarks in RoiSets (.zip), measurements in tabular format (.txt)
- `2-plots`: Plotted data from ROIs

## Pre-processing

We exported the measurements of individual embryos using the "Measure" command
from Fiji. Each file contains 4 rows with predefined values. Row 1 and 2 are the
anterior and posterior limits of the vitelline envelope, respectively. The Slice
value of these two rows represents the frame at the onset of gastrulation (T=0).
Row 3 marks the position of the tip of the germ band when the cephalic furrow or
an ectopic fold first buckle in. The Slice correspond to the frame when this
happens. Row 4 marks the position of the tip of the germ band when the cephalic
furrow or ectopic folds reach maximum depth, immediately before they begin to
unfold. The Slice also corresponds to the frame when this happens.

Here is a sample file:

```
 	Label	X	Y	Slice
1	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	72.000	358.000	26
2	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	1820.833	314.167	26
3	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	1240.833	119.500	47
4	3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55	1050.167	115.500	56
```

For extracting information and plotting, these raw files need to be parsed,
processed, and compiled in a single data frame.

```{r preprocess}
# List all txt files in ROIs folder
# Function to get and process data from individual files imported as dataframes
# Loop to iterate and actually process the data
# Use row bind to build a novel data frame with the compiled data
```




