---
title: "Germ band extension relative to ectopic folding"
author: "Bruno C. Vellutini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
creation: October 24, 2022
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/Projects/Cephalic_furrow/7-Manuscript/1-analyses/germband-extension")
library(ggplot2)
library(ggbeeswarm)
```

## Pre-processing

We exported the measurements of individual embryos using the "Measure" command from Fiji. Each file contains 4 rows with predefined values. Row 1 and 2 are the anterior and posterior limits of the vitelline envelope, respectively. The Slice value of these two rows represents the frame at the onset of gastrulation (T=0). Row 3 marks the position of the tip of the germ band when the cephalic furrow or an ectopic fold first buckle in. The Slice correspond to the frame when this happens. Row 4 marks the position of the tip of the germ band when the cephalic furrow or ectopic folds reach maximum depth, immediately before they begin to unfold. The Slice also corresponds to the frame when this happens.

1.  anterior_end (frame_gastrulation)
2.  posterior_end (frame_gastrulation)
3.  first_buckle (CF or EF)
4.  max_depth (CF or EF)

Here is a sample file:

```         
    Label   X   Y   Slice
1   3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55    72.000  358.000 26
2   3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55    1820.833    314.167 26
3   3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55    1240.833    119.500 47
4   3D_btd-gap2_z3_t55s_E7.tif:0056-0237-0946:55    1050.167    115.500 56
```

For extracting information and plotting, these raw files need to be parsed, processed, and compiled in a single data frame. The next block reads, process, and creates an initial data frame.

```{r preprocess-fold-start}
# List all txt files in ROIs folder
files = list.files(path = "1-rois/fold_start", pattern="*.tsv", full.names = TRUE)

# Function for pre-processing
process_fold <- function(txtfile) {
  # Read txt file
  df <- read.delim(txtfile)
  # Get label and clean up embryo name
  label_tf <- df$Label[1]
  embryo <- strsplit(df$Label[1], ":")[[1]][1]
  embryo <- gsub("3D_", "", embryo)
  embryo <- gsub(".tif", "", embryo)
  # Get X positions
  x_ant <- df$X[1]
  x_post <- df$X[2]
  x_tf_buck <- df$X[3]
  x_tf_deep <- df$X[4]
  # Get sizes
  size <- x_post - x_ant
  gb_pos_tf_buck <- x_tf_buck - x_ant
  gb_pos_tf_deep <- x_tf_deep - x_ant
  # Get percentages
  gb_pct_tf_buck <- 100 - ((gb_pos_tf_buck / size) * 100)
  gb_pct_tf_deep <- 100 - ((gb_pos_tf_deep / size) * 100)
  # Get frames
  frame_gastr <- df$Slice[1]
  frame_tf_buck <- df$Slice[3]
  frame_tf_deep <- df$Slice[4]
  # Save as vector
  mv <- c(label_tf, embryo, size, x_ant, x_post, x_tf_buck, x_tf_deep,
          gb_pos_tf_buck, gb_pos_tf_deep, gb_pct_tf_buck, gb_pct_tf_deep,
          frame_gastr, frame_tf_buck, frame_tf_deep)
  return(mv)
}

# Define empty data frame
fold_start <- data.frame()

# Loop over files binding rows to data frame
for (txtfile in files) {
  gbrow <- process_fold(txtfile)
  fold_start <- rbind(fold_start, gbrow)
}

# Define column names
colnames(fold_start) <- c("label_tf", "embryo", "size", "x_ant", "x_post", "x_tf_buck", "x_tf_deep",
                         "gb_pos_tf_buck", "gb_pos_tf_deep", "gb_pct_tf_buck", "gb_pct_tf_deep",
                         "frame_gastr", "frame_tf_buck", "frame_tf_deep")

# Save data frame to file
write.csv(fold_start, file = "gb_extension.csv", row.names = FALSE)
```

```{r preprocess-md-start}
# List all txt files in ROIs folder
files = list.files(path = "1-rois/md_start", pattern="*.tsv", full.names = TRUE)

# Function for pre-processing
process_md <- function(txtfile) {
  # Read txt file
  df <- read.delim(txtfile)
  # Get label and clean up embryo name
  label_md <- df$Label[1]
  embryo <- strsplit(df$Label[1], ":")[[1]][1]
  embryo <- gsub("3D_", "", embryo)
  embryo <- gsub(".tif", "", embryo)
  # Get X positions
  x_md_start <- df$X[1]
  x_md_split <- df$X[2]
  # TODO: Get sizes
  # TODO: Get percentages
  
  # Get frames
  frame_md_start <- df$Slice[1]
  frame_md_split <- df$Slice[2]

    # Save as vector
  mv <- c(label_md, embryo, x_md_start, x_md_split, frame_md_start, frame_md_split)
  return(mv)
}

# Define empty data frame
md_start <- data.frame()

# Loop over files binding rows to data frame
for (txtfile in files) {
  gbrow <- process_md(txtfile)
  md_start <- rbind(md_start, gbrow)
}

# Define column names
colnames(md_start) <- c("label_md", "embryo", "x_md_start", "x_md_split", "frame_md_start", "frame_md_split")
```

## Dataset building

Next, we need to import the master dataset file to get some metadata. Most important is the stock, zygosity, timestep, and gastrulation frame.

```{r merge}
# Import main dataset
datasets <- read.csv("../../0-data/datasets.csv")

# Merge landmarks with datasets
germband <- merge(fold_start, datasets, by = "embryo", all.x = TRUE)
germband <- merge(germband, md_start, by = "embryo", all.x = TRUE)

# Properly convert column types before any calculations
germband <- as.data.frame(lapply(germband, type.convert))

# Calculate GB percentage for MD start
germband$gb_pos_md_start <- germband$x_md_start - germband$x_ant
germband$gb_pos_md_split <- germband$x_md_split - germband$x_ant
germband$gb_pct_md_start <- 100 - ((germband$gb_pos_md_start / germband$size) * 100)
germband$gb_pct_md_split <- 100 - ((germband$gb_pos_md_split / germband$size) * 100)

# Create columns with times (min)
germband$time_tf_buck <- ((germband$frame_tf_buck - germband$frame_gastr) * germband$timestep) / 60
germband$time_tf_deep <- ((germband$frame_tf_deep - germband$frame_gastr) * germband$timestep) / 60
germband$time_md_start <- ((germband$frame_md_start - germband$frame_gastr) * germband$timestep) / 60
germband$time_md_split <- ((germband$frame_md_split - germband$frame_gastr) * germband$timestep) / 60

```

Convert the main dataframe into the long format for plotting

```{r gblong}
# Define variables for plotting
longvars <- c(
  "embryo",
  "stock",
  "zygosity",
  "size",
  "gb_pct_tf_buck",
  "gb_pct_tf_deep",
  "gb_pct_md_start",
  "gb_pct_md_split",
  "time_tf_buck",
  "time_tf_deep",
  "time_md_start",
  "time_md_split"
)

gblong <- reshape(
  germband[, longvars],
  varying = list(
    c(
      "gb_pct_tf_buck",
      "gb_pct_tf_deep",
      "gb_pct_md_start",
      "gb_pct_md_split"
    ),
    c(
      "time_tf_buck",
      "time_tf_deep",
      "time_md_start",
      "time_md_split"
    )
  ),
  v.names = c("gb_pct", "time"),
  timevar = "status",
  times = c("tf_buck", "tf_deep", "md_start", "md_split"),
  direction = "long"
)

# Remove the id column that reshape adds
gblong$id <- NULL

# Transform data frame to order the stock for plotting
gblong$stock = factor(gblong$stock, levels = c("wt", "btd", "eve", "prd", "stg", "btd-stg"))
gblong$zygosity = factor(gblong$zygosity, levels = c("wildtype", "heterozygote", "homozygote", "double heterozygote", "btd homozygote", "stg homozygote", "double homozygote"))
gblong$status = factor(gblong$status, levels = c("tf_buck", "tf_deep", "md_start", "md_split"))
gblong <- gblong[order(gblong$stock, gblong$zygosity),]
rownames(gblong) <- 1:nrow(gblong)

# Save both data frames to file
write.csv(germband, file = "germband.csv", row.names = FALSE)
write.csv(gblong, file = "gblong.csv", row.names = FALSE)
```

## Descriptive statistics

Calculate and print the mean and standard deviation for all variables.

```{r values}
stats <- aggregate(cbind(gb_pct, time) ~ stock + zygosity + status, gblong, function(x) c(count = length(x), mean = mean(x), sd = sd(x)))
stats <- do.call(data.frame, stats)
stats <- stats[order(stats$status, stats$stock, stats$zygosity),]
rownames(stats) <- 1:nrow(stats)
stats
write.csv(stats, file = "stats.csv", row.names = FALSE)
```

## Descriptive plots

We want to identify the difference between the cephalic furrow and the ectopic folds in the percentage of germ band extension, and the timing of formation after gastrulation.

First, we plot the germ band extension (%) for homozygotes and heterozygotes at two different moments: initial infolding/buckling (buck) and deepest point (deep).

```{r plot-gb-pct}
ggplot(gblong, aes(stock, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "germ band extension (%)") +
  facet_grid(rows = vars(status), cols = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct.png", width = 15, height = 12)
ggsave("2-plots/gb_pct.svg", width = 15, height = 12)
```

Second, we plot the time after gastrulation (min).

```{r plot-gb-time}
ggplot(gblong, aes(stock, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = TRUE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = TRUE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "time after gastrulation (min)") +
  facet_grid(rows = vars(status), cols = vars(stock), scales = "free_x") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        title = element_text(size = 15),
        legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time.png", width = 15, height = 12)
ggsave("2-plots/gb_time.svg", width = 15, height = 12)
```

## Main figure plots

The above are nice descriptive plots. But for the figure, the most interesting is the initial buckling/infolding relative to GB position and time after gastrulation.

Plot germ band extension percentage at the moment of buckling/infolding:

```{r plot-gb-pct-buck}
ggplot(gblong[gblong$status == "tf_buck",], aes(stock, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "germ band extension (%)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_buck.png", width = 12, height = 4)
ggsave("2-plots/gb_pct_buck.svg", width = 12, height = 4)
```

Plot the time after gastrulation at the moment of buckling/infolding:

```{r plot-gb-time-buck}
ggplot(gblong[gblong$status == "tf_buck",], aes(stock, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_buck.png", width = 12, height = 4)
ggsave("2-plots/gb_time_buck.svg", width = 12, height = 4)
```

Plot the germ band extension at the time of buckling/infolding and mitotic domain expansion:

```{r plot-gb-time-buck}
ggplot(gblong[(gblong$status %in% c("tf_buck", "md_start") & gblong$stock != "prd"),], aes(status, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  #geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE) +
  geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "germ band extension (%)") +
  facet_grid(cols = vars(zygosity), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_tf_buck_md_start.png", width = 9, height = 4)
ggsave("2-plots/gb_pct_tf_buck_md_start.svg", width = 9, height = 4)
```

Plot the time after gastrulation at the time of buckling/infolding and mitotic domain expansion:

```{r plot-gb-time-buck}
ggplot(gblong[(gblong$status %in% c("tf_buck", "md_start") & gblong$stock != "prd"),], aes(status, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  #geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE) +
  geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(zygosity), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_tf_buck_md_start.png", width = 9, height = 4)
ggsave("2-plots/gb_time_tf_buck_md_start.svg", width = 9, height = 4)
```

Plot the germ band percentage vs buckling/infolding and md expansion, but faceted by status:

```{r plot-gb-time-buck}
ggplot(gblong[(gblong$status %in% c("tf_buck", "md_start") & gblong$stock != "prd"),], aes(zygosity, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 2, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  #geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  #geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "zygosity", y = "germ band extension (%)") +
  facet_grid(cols = vars(status), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_tf_buck_md_start_facet_status.png", width = 6, height = 4)
ggsave("2-plots/gb_pct_tf_buck_md_start_facet_status.svg", width = 6, height = 4)
```

Plot the time after gastrulation vs buckling/infolding and md expansion, but faceted by status:

```{r plot-gb-time-buck}
ggplot(gblong[(gblong$status %in% c("tf_buck", "md_start") & gblong$stock != "prd"),], aes(zygosity, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 2, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 4)) +
  #geom_point(alpha = 0.8, size = 3, show.legend = TRUE) +
  #geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "zygosity", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(status), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_tf_buck_md_start_facet_status.png", width = 6, height = 4)
ggsave("2-plots/gb_time_tf_buck_md_start_facet_status.svg", width = 6, height = 4)
```

## Supplementary plots

The deepest point plots are not super informative and will go into the supplementary.

```{r plot-gb-pct-deep}
ggplot(gblong[gblong$status == "tf_deep",], aes(stock, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "germ band extension (%)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        #title = element_text(size = 15),
        #legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_deep.png", width = 12, height = 4)
ggsave("2-plots/gb_pct_deep.svg", width = 12, height = 4)
```

```{r plot-gb-time-deep}
ggplot(gblong[gblong$status == "tf_deep",], aes(stock, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "stocks", y = "time after gastrulation (min)") +
  facet_grid(cols = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        #title = element_text(size = 15),
        #legend.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_deep.png", width = 12, height = 4)
ggsave("2-plots/gb_time_deep.svg", width = 12, height = 4)
```

```{r plot-gb-pct-tf-buck-md-start-by-stock}
ggplot(gblong[gblong$status %in% c("tf_buck", "md_start"),], aes(status, gb_pct, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  #geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "germ band extension (%)") +
  #facet_grid(cols = vars(zygosity), scales = "free") +
  facet_grid(cols = vars(status), rows = vars(stock), scales = "free_x") +
  #facet_grid(rows = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_pct_tf_buck_md_start_by_stock.png", width = 7, height = 12)
ggsave("2-plots/gb_pct_tf_buck_md_start_by_stock.svg", width = 7, height = 12)
```

```{r plot-gb-time-tf-buck-md-start-by-stock}
ggplot(gblong[gblong$status %in% c("tf_buck", "md_start"),], aes(status, time, fill = zygosity)) +
  geom_boxplot(aes(color = zygosity), show.legend = FALSE, alpha = 0.5, outlier.shape = NA, width = 0.7, position = position_dodge(width = 1.0)) +
  geom_point(alpha = 0.8, size = 3, show.legend = FALSE, position = position_beeswarm(priority = "descending", dodge.width = 1.0, cex = 6)) +
  #geom_line(aes(group = embryo)) +
  scale_fill_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  scale_color_manual(values=c("#cc79a7", "#56b4e9", "#e69f03")) +
  labs(x = "status", y = "time after gastrulation (min)") +
  #facet_grid(cols = vars(zygosity), scales = "free") +
  facet_grid(cols = vars(status), rows = vars(stock), scales = "free_x") +
  #facet_grid(rows = vars(stock), scales = "free") +
  theme_classic(base_family = "DejaVu Sans") +
  theme(text = element_text(size = 35),
        strip.background = element_blank(),
        panel.background = element_rect(colour = "black", size=1)
        )
ggsave("2-plots/gb_time_tf_buck_md_start_by_stock.png", width = 7, height = 12)
ggsave("2-plots/gb_time_tf_buck_md_start_by_stock.svg", width = 7, height = 12)
```

## Statistical tests

```{r tests}
# Create data frame for tests
tests <- data.frame()

# Contrasts of single variables within stocks (het vs hom)
for (stock in unique(gblong$stock)) {
  
  if (stock == "wt") { next }
  
  for (status in unique(gblong$status)) {
    
    for (value in c("gb_pct", "time")) {
      
      het = gblong[(gblong$status == status & gblong$stock == stock & gblong$zygosity == "heterozygote"), value]
      hom = gblong[(gblong$status == status & gblong$stock == stock & gblong$zygosity == "homozygote"), value]
      
      test = wilcox.test(het, hom, paired = FALSE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(het) - mean(hom)
      
      test_row = c(
        contrast = "het vs hom",
        stock = stock,
        zygosity = "-",
        status = status,
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.6f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = FALSE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
    }
  }
}

# Contrasts between variables within zygosity (buck vs md)
for (stock in unique(gblong$stock)) {
  
  for (zygosity in unique(gblong$zygosity)) {
    
    for (value in c("gb_pct", "time")) {
      
      if (stock == "wt" & zygosity != "wildtype") { next }
      
      if (stock != "wt" & zygosity == "wildtype") { next }
      
      # Note: with low n and all values from one contrast higher or lower than the other, pvalues might be identical
      buck = gblong[(gblong$status == "tf_buck" & gblong$stock == stock & gblong$zygosity == zygosity), value]
      md = gblong[(gblong$status == "md_start" & gblong$stock == stock & gblong$zygosity == zygosity), value]
      
      test = wilcox.test(buck, md, paired = TRUE)
      pvalue = test$p.value
      statistic = test$statistic
      delta = mean(buck) - mean(md)
      
      test_row = c(
        contrast = "buck vs md",
        stock = stock,
        zygosity = zygosity,
        status = "-",
        value = value,
        delta = sprintf("%.1f", delta),
        significant = pvalue < 0.05,
        pvalue = sprintf("%.6f", pvalue),
        statistic = sprintf("%s=%s", names(statistic), statistic),
        paired = TRUE
      )
      
      tests <- rbind(tests, as.data.frame(t(test_row)))
      
    }
  }
}

# Contrasts between variables within zygosity but batch stocks (buck vs md)
for (zygosity in c("heterozygote", "homozygote")) {
  
  for (value in c("gb_pct", "time")) {
    
    # Note: prd excluded because CF formation is not prevented
    buck = gblong[(gblong$status == "tf_buck" & gblong$stock != "prd" & gblong$zygosity == zygosity), value]
    md = gblong[(gblong$status == "md_start" & gblong$stock != "prd" & gblong$zygosity == zygosity), value]
    
    test = wilcox.test(buck, md, paired = TRUE)
    pvalue = test$p.value
    statistic = test$statistic
    delta = mean(buck) - mean(md)
    
    test_row = c(
      contrast = "buck vs md",
      stock = "btd+eve",
      zygosity = zygosity,
      status = "-",
      value = value,
      delta = sprintf("%.1f", delta),
      significant = pvalue < 0.05,
      pvalue = sprintf("%.6f", pvalue),
      statistic = sprintf("%s=%s", names(statistic), statistic),
      paired = TRUE
    )
    
    tests <- rbind(tests, as.data.frame(t(test_row)))
  }
}

tests
write.csv(tests, file = "tests.csv", row.names = FALSE)
```
